---
title: "Performance of linking researchers to theses"
author: "Flavio & Christoph"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
---

This script makes some plots of the advisor links and saves the most plausible links to a table in the database. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

```{r, echo = FALSE, include = FALSE}
packages <- c("tidyverse", "broom", "dbplyr", "RSQLite", "ggplot2", "stringdist")

lapply(packages, library, character.only = TRUE)

datapath <- "/mnt/ssd/"
db_file  <- paste0(datapath, "AcademicGraph/AcademicGraph.sqlite")
select_fields <- c("physics", "biology", "chemistry", "sociology",
                    "economics", "political science", "psychology", 
                    "mathematics", "geography", "geology", "engineering",
                    "computer science", "environmental science") # fields currently matched 


# ## db connection
con <- DBI::dbConnect(RSQLite::SQLite(), db_file)
cat("The database connection is: \n")
src_dbi(con)
```



```{r}
# parameters for selecting links 
min_score_advisors <- 0.7 # minimum score from dedupe 
max_year_diff <- 5 # maximum difference between advisory and own publication at institution. 5 is arbitrary but perhaps makes sense: I publish something at Aalto, start advising a first-year student, then move elsewhere. then I would still figure as advisor at Aalto
max_uniname_distance <- 0.02 # keep only links where the jarowinkler distance between the institution naems is below that threshold. 0.02 is conservative


```


## Overview

```{r, include = FALSE}

current_links <- tbl(con, sql("
                              select * 
                              from current_links 
                              inner join (
                                select authorid, normalizedname as author_name
                                from authors 
                              ) using(authorid)
                              inner join (
                                select AuthorId, main_institutions_career
                                from author_info_linking
                              ) using(authorId)
                              "))
linking_info <- tbl(con,
                    sql("select iteration_id, field
                          from (
                              select *, max(iteration_id) OVER(partition by field) as max_iter
                              from linking_info_advisors
                              where recall = 0.9 
                                  and testing = 0
                                  and institution = 'True'
                                  and fieldofstudy_cat = 'False'
                                  and fieldofstudy_str = 'False' 
                                  and keywords = 'False' 
                                  and mergemode = 'm:1'
                          )
                          where iteration_id = max_iter "))
linked_advisors <- tbl(con, sql("select * 
                                from linked_ids_advisors
                                ")) %>%
  inner_join(linking_info %>% 
               select(iteration_id),
             by = "iteration_id")



theses <- tbl(con, sql("
                       select * from pq_advisors 
                       inner join (
                         select goid, degree_year, uni_name 
                         from pq_authors
                         inner join (
                          select university_id, normalizedname as uni_name 
                          from pq_unis
                         ) using(university_id)
                       ) using(goid)")
              )

authors_affiliation <- tbl(con, 
                           sql("
                               select authorid, year, affil_name
                               from AuthorAffiliation
                               inner join (
                                select affiliationid, normalizedname as affil_name
                                from affiliations
                               ) using(affiliationid)
                               inner join (
                                select authorid
                                from current_links
                               ) using(authorid)
                               "))

```


```{r}

current_links <- collect(current_links)
linked_advisors <- collect(linked_advisors)
theses <- collect(theses)
authors_affiliation <- collect(authors_affiliation)
linking_info <- collect(linking_info)
```


## Linking scores 

- conditioning on link score > 0.7 is fine 

```{r}

linked_advisors %>%
  left_join(linking_info, by = "iteration_id") %>%
  ggplot(aes(x = link_score)) +
  geom_histogram(bins = 100, aes( y = ..density..)) +
  facet_wrap(~field)

```

## Average link score by graduation year


```{r}

score_by_year <- linked_advisors %>%
  left_join(theses %>%
              select(degree_year, relationship_id),
            by = "relationship_id") %>%
  left_join(linking_info,
            by = "iteration_id")

score_by_year %>%
  group_by(degree_year, field) %>%
  summarise(mean_score = mean(link_score),
            p50_score = quantile(link_score, probs = 0.5),
            .groups = "drop") %>%
  pivot_longer(cols = ends_with("_score"),
               names_to = "stat") %>%
  ggplot(aes(x = degree_year, y = value)) + 
  geom_line(aes(linetype = stat)) + 
  facet_wrap(~field) + 
  theme(legend.position = "bottom")


score_by_year %>%
  group_by(degree_year, field) %>%
  summarise(n_advisors = n(),
            .groups = "drop") %>%
  ggplot(aes(x = degree_year, y = n_advisors)) + 
  geom_line() + 
  facet_wrap(~field)


# another stat
  # linked advisors per linked dissertation
  # versus advisors per dissertation (in the US)

# score_by_year %>% 
#   mutate(grp = ifelse(degree_year < 2000, "pre", "post")) %>%
#   ggplot(aes(x = link_score)) +
#   geom_histogram() + 
#   facet_wrap(~grp)


```

old comments

- for instance, a student of michael j lambert (authorid 2120159045; relationship id 303670971_0 in proquest) from pre-1990 is link score of 0.02, but should be a clear link


## Compare number of links across iterations within fields 

```{r}
fields_iter_compare <- c("economics", "chemistry")
min_score <- 0.8

keep_iter_ids <- tbl(con, "linking_info_advisors") %>%
  filter(field %in% fields_iter_compare) %>%
  filter(testing == 0) %>%
  collect() %>%
  group_by(field, train_name) %>%
  arrange(iteration_id) %>%
  mutate(nb = n(),
         id = row_number()) %>%
  ungroup() %>%
  filter(id == nb) %>%
  select(iteration_id, field, train_name)

linked_ids_to_compare <- tbl(con, "linked_ids_advisors") %>%
  inner_join(
    tbl(con, "linking_info_advisors") %>%
      filter(field %in% fields_iter_compare),
    by = "iteration_id"
  ) %>%
  inner_join(
    tbl(con, "pq_advisors") %>%
      select(relationship_id, goid),
    by = "relationship_id"
  ) %>%
  inner_join(
    tbl(con, "pq_authors") %>%
      select(goid, degree_year),
    by = "goid"
  ) %>%
  collect() %>%
  filter(iteration_id %in% keep_iter_ids$iteration_id) 

```


Number of graduates with at least 1 advisor

```{r}

d_sum <- linked_ids_to_compare %>%
  filter(link_score >= min_score) %>%
  group_by(train_name, field, degree_year) %>%
  summarise(n_advisors = n(),
            n_graduates = n_distinct(goid),
            .groups = "drop") %>%
  pivot_longer(cols = starts_with("n_"), names_to = "variable")

plotvars <- c("n_graduates")

map(.x = plotvars,
    .f = ~d_sum %>%
      filter(variable == .x) %>%
      ggplot(aes(x = degree_year, y = value)) +
      geom_line(aes(linetype = train_name)) + 
      facet_wrap(~field) + 
      theme(legend.position = "bottom") + 
      labs(title = paste0("Count: ", .x))
    )


```





## Check overlap of institution names and years 

```{r}

d_main <- linked_advisors %>%
  filter(link_score > min_score_advisors) %>%
  left_join(theses %>%
              mutate(fullname = paste0(firstname, " ", lastname)) %>%
              select(relationship_id, degree_year, uni_name, fullname),
            by = "relationship_id") %>%
  inner_join(current_links %>%
               select(author_name, AuthorId),
             by = "AuthorId") %>%
  # join on year; filter on max similarity within relationship_id; still need to examine multiple matches (relationship_id) later; compare the years (they may not match if the author moves on but still has advisees at the old place )
  left_join(authors_affiliation, 
            by = c("AuthorId")) %>%
  mutate(dist_uni_name = stringdist(uni_name, affil_name, method = "jw"),
         dist_year = abs(degree_year - Year)) %>%
  group_by(relationship_id) %>%
  filter(dist_uni_name == min(dist_uni_name)) %>%
  filter(dist_year == min(dist_year)) %>%
  mutate(nb = n()) %>% # can still have multiple links if e.g. the dissertation is in x, but the affiliation is in x-1 and x+1 at the same institution
  ungroup() 

d_main <- d_main %>% 
  filter(!duplicated(relationship_id))

cat("Split of links by whether years are >/<", max_year_diff, "apart")
table(d_main$dist_year <= max_year_diff)


d_main <- d_main %>%
  filter(dist_year <= max_year_diff & dist_uni_name <= max_uniname_distance)


```

## Note: the "usable" links are saved to the db in src/dataprep/main/link/prep_linked_data.py




## Fraction with matched advisor status by cohort of own graduation 

```{r}
pq_authors <- tbl(con, "pq_authors") %>% collect()

```


```{r}

d_agg <- d_main %>%
  select(AuthorId, relationship_id, degree_year) %>%
  group_by(AuthorId) %>%
  filter(degree_year == min(degree_year)) %>%
  ungroup() %>%
  rename(year_firstadvisee = degree_year) %>%
  filter(!duplicated(AuthorId))

```


```{r}
d_links <- current_links %>%
  left_join(pq_authors %>%
              select(goid, year_phd = degree_year),
            by = "goid") %>%
  left_join(d_agg, by = "AuthorId")


```


Fraction of authors that eventually becomes advisor
```{r}

d_links %>%
  mutate(advisor = ifelse(is.na(year_firstadvisee), 0, 1)) %>%
  group_by(year_phd) %>%
  summarise(advisor = mean(advisor),
            .groups = "drop") %>%
  ggplot(aes(x = year_phd, y = advisor)) +
  geom_point() + 
  geom_line()

```

Duration to advisor

```{r}

d_links %>%
  filter(!is.na(year_firstadvisee)) %>%
  mutate(duration = year_firstadvisee - year_phd) %>%
  group_by(year_phd) %>%
  summarise(duration = mean(duration),
            .groups = "drop") %>%
  ggplot(aes(x = year_phd, y = duration)) +
  geom_point() + 
  geom_line()


```



```{r, include = FALSE}
DBI::dbDisconnect(con)
```
