---
title: "Compare Linking across linking runs"
author: "Christoph, Flavio, Mona"
date: "2023-07-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
datapath <- "/mnt/ssd/linked_ids_temp/"
packages <- c("tidyverse", "broom", "dbplyr", "RSQLite", "ggplot2", 
              "stringdist", "knitr", "bit64", "kableExtra",
              "magutils") 

lapply(packages, library, character.only = TRUE)

datapath <- "/mnt/ssd/"
db_file  <- paste0(datapath, "AcademicGraph/AcademicGraph.sqlite")
con <- DBI::dbConnect(RSQLite::SQLite(), db_file)
cat("The database connection is: \n")
src_dbi(con)

```

## Linking Advisors and Graduates from ProQuest to MAG AuthorIds

To check whether the linking makes sense we compare the links across several independent runs of the labelling.

```{r comparison function, include = FALSE}

compare <- function(field, linktype, linker1, linker2, years, inspect=FALSE){
  
  if (linktype=="advisors"){
      col_names1 = c("pqid","AuthorId_1","link_score_1")
      col_names2 = c("pqid","AuthorId_2","link_score_2")
  } else {
      col_names1 = c("AuthorId_1","pqid","link_score_1")
      col_names2 = c("AuthorId_2","pqid","link_score_2") 
  }
  
  file1 = paste0(datapath,"linked_ids_temp/links_",linktype,"_", field, "_",linker1,"_",years,".csv")
  file2 = paste0(datapath,"linked_ids_temp/links_",linktype,"_", field, "_",linker2,"_",years,".csv")  
  file1 <- ifelse(file.exists(file1), file1, 
                 paste0(datapath,"linked_ids_temp/links_",linktype,"_", field, "_",str_replace(linker1, "_degree0", ""),"_fielddegree0.csv"))
  file2 <- ifelse(file.exists(file2), file2, 
                 paste0(datapath,"linked_ids_temp/links_",linktype,"_", field, "_",str_replace(linker2, "_degree0", ""),"_fielddegree0.csv"))

  
  links_1 <- read_csv(file1, 
                      skip=1, 
                      col_names=col_names1,
                      col_types=list(col_character(), 
                                     col_character(), 
                                     col_double())) %>% filter(link_score_1>0.7)
  

  links_2 <- read_csv(file2, 
                      skip=1, 
                      col_names=col_names2,
                      col_types=list(col_character(), 
                                     col_character(), 
                                     col_double()))%>% filter(link_score_2>0.7)

 
  
  links <- full_join(links_1, links_2, by="pqid") 
  
  links <- links %>% mutate(
                  samelink = (AuthorId_1==AuthorId_2)&!is.na(link_score_1),
                  link1 = !is.na(link_score_1),
                  link2 = !is.na(link_score_2),
                  link_only1 = !is.na(link_score_1)&is.na(link_score_2),
                  link_only2 = !is.na(link_score_2)&is.na(link_score_1),
                  bothlink = !is.na(link_score_1)&!is.na(link_score_2),
                  difflink = (AuthorId_1!=AuthorId_2)&bothlink
                  ) # check that NA is correct or if "" is in there
  
  
  ids1 <- links %>% select(AuthorId=AuthorId_1) %>% drop_na() %>% pull(AuthorId)
  ids2 <- links %>% select(AuthorId=AuthorId_2) %>% drop_na() %>% pull(AuthorId)
  ids <- union(ids1, ids2)  

 # load NormalizedName of AuthorId_1 and AuthorId_2
  query <- paste0(
    "SELECT AuthorId, NormalizedName, PaperCount 
    FROM Authors
    WHERE AuthorId IN (", 
    paste0(ids, collapse = ","),
    ") "
  )
  
  author_names <- tbl(con, sql(query)) |>
    collect() |>
    mutate(AuthorId = as.character(AuthorId))


  links <- author_names %>% 
    right_join(links, by=c("AuthorId" = "AuthorId_1")) %>% 
    rename(NormalizedName_1=NormalizedName, PaperCount_1 = PaperCount) %>% 
    rename(AuthorId_1=AuthorId)
  
  links <- author_names %>% 
   right_join(links, by=c("AuthorId" = "AuthorId_2")) %>% 
   rename(NormalizedName_2=NormalizedName, PaperCount_2 = PaperCount) %>%  
   rename(AuthorId_2=AuthorId)
  if (linktype=="advisors"){
  # load name from PQ
  pq_ids <- links |> pull(pqid) |> unique()
  query <- paste0(
    "SELECT relationship_id as pqid
     , firstname || ' ' || lastname as normalizedname_pq
    FROM pq_advisors 
    WHERE relationship_id IN (",
    paste0(paste0("'", pq_ids, "'"), collapse = ","),
    ")"
  )
  } else {
  # load name from PQ
  pq_ids <- links |> pull(pqid) |> unique()
  query <- paste0(
    "SELECT goid as pqid
     , fullname as normalizedname_pq
    FROM pq_authors 
    WHERE goid IN (",
    paste0(paste0("'", pq_ids, "'"), collapse = ","),
    ")"
  )
  }
  pq_names <- tbl(con, sql(query)) |>
    collect() %>% mutate(pqid=as.character(pqid))

  links <- links |> 
    left_join(pq_names, by = "pqid") 
  links <- links %>% 
    mutate(namedist = stringdist(NormalizedName_1, NormalizedName_2, method="jw", p=0),
           namedist_pq_1 = stringdist(NormalizedName_1, normalizedname_pq, method = "jw", p=0),
           namedist_pq_2 = stringdist(NormalizedName_2, normalizedname_pq, method = "jw", p=0),
           namedist_lv = stringdist(NormalizedName_1, NormalizedName_2, method="lv"),
           namedist_lv_pq_1 = stringdist(NormalizedName_1, normalizedname_pq, method = "lv"),
           namedist_lv_pq_2 = stringdist(NormalizedName_2, normalizedname_pq, method = "lv"),
           namedist_qgram = stringdist(NormalizedName_1, NormalizedName_2, method="qgram"),
           namedist_qgram_pq_1 = stringdist(NormalizedName_1, normalizedname_pq, method = "qgram"),
           namedist_qgram_pq_2 = stringdist(NormalizedName_2, normalizedname_pq, method = "qgram"))
 
  shares <- links %>% summarize(same = sum(samelink, na.rm=TRUE)/n_distinct(pqid),
                                only1 = mean(link_only1),
                                only2 = mean(link_only2),
                                diff = mean(difflink),
                                diff_rel1 = sum(difflink*link1)/sum(link1),
                                nlink1 = n_distinct(AuthorId_1),
                                nlink2 = n_distinct(AuthorId_2),
                                namedist_pq_1 = sum(namedist_pq_1,
                                                         na.rm=TRUE)/sum(link1),
                                namedist_pq_2 = sum(namedist_pq_2,
                                                         na.rm=TRUE)/sum(link2),
                                namedist_diff = sum(namedist * difflink, na.rm=TRUE)/sum(difflink),
                                namedist_diff_pq_1 = sum(namedist_pq_1 * difflink,
                                                         na.rm=TRUE)/sum(difflink),
                                namedist_diff_pq_2 = sum(namedist_pq_2 * difflink,
                                                         na.rm=TRUE)/sum(difflink),
                                namedist_pq_only1 = sum(namedist_pq_1 * link_only1,
                                                        na.rm=TRUE)/sum(link_only1*!is.na(namedist_pq_1)),
                                namedist_pq_only2 = sum(namedist_pq_2 * link_only1,
                                                        na.rm=TRUE)/sum(link_only2*!is.na(namedist_pq_2))
                                )
  shares$field = field
  shares <- shares %>% relocate(field)
  #print(paste("Field", field, " linktype ", linktype, "comparison for ", linker1, "and", linker2))
  #print(shares)
  if (inspect){
    return(links)
  } else {
  return(shares)
  }
}
 
```


### Which fields and which linking runs to compare?


```{r comparison setup graduates, include=TRUE}
fields_to_process <-c( #"art", <- dropped
  "biology", 
  "business",
  "chemistry",
  "computer science" ,
  "economics",
  "engineering",
  "environmental science", 
  "geography", 
  "geology" ,
  "history",
  "materials science",
  "mathematics",
  #"medicine",   <- dropped
  "philosophy",
  "physics",
  "political science",
  "psychology" ,
  "sociology")

linker1 = "christoph_with_protocol_magkeywords"
linker2 = "flavio_with_protocol_magkeywords"
```



# Load additional data from proquest for tables

```{r graduate: N theses, N links, include=TRUE}
all_graduates <- get_proquest(conn = con, from = "graduates", start_year = 1990, end_year = 2015) |>
  collect()

graduate_links <- get_links(conn = con, from = "graduates") |> 
  collect() |> 
  filter(goid %in% all_graduates$goid)


advisor_links <- get_links(conn = con, from = "advisors") |>
  left_join(tbl(con, "pq_advisors") |>
              select(goid, relationship_id),
            by = "relationship_id") |> 
  collect() |>
  filter(goid %in% all_graduates$goid) 

d_links <- list(
  graduates = graduate_links,
  advisors = advisor_links
)

d_links <- map(
  .x = d_links,
  .f = ~.x |> 
    left_join(all_graduates |> 
                select(goid, fieldname0_mag),
              by = "goid")
)


# some are not loaded in get_proquest b/c they have special degrees (Psy.D). -> filter on goid being in all_graduates 
# others are in get_proquest but have missing field of study. why? -- but they need to have a link in the first place!


```


If some linked graduates in `get_proquest` had missing fields, it would show up here and should be addressed.

```{r, include=TRUE}
map(d_links, ~mean(is.na(.x[["fieldname0_mag"]])))

map(d_links, ~.x |> filter(is.na(fieldname0_mag)) |> head())

# count is zero for advisors: this means that none of the graduates with missing field above are in the linked, which is what I expect
# Note: a complication is that linked graduates and all_graduates both have units that are not in the other set
  # (this is because the sampling for linking is different than the sampling in get_proquest)
# therefore, it's also important to make the sets overlapping, and then compare the links found/check whether they have fields missing
```



```{r numbers by field, include=TRUE}

graduate_counts <- list(
  "total" = all_graduates,
  "links" = d_links$graduates
)

graduate_counts <- map(
  graduate_counts,
  ~ .x |> 
    filter(fieldname0_mag %in% fields_to_process) |> 
    group_by(fieldname0_mag) |> 
    summarise(nb = n())
)

advisor_link_counts <- d_links$advisors |>
  filter(fieldname0_mag %in% fields_to_process) |> 
  group_by(fieldname0_mag) |> 
  summarise(nb = n())
```


# Comparison for Graduates

```{r comparison graduates, include=FALSE}

# Loop through the fields
res <- lapply(fields_to_process, function(x) compare(x,  "graduates", linker1, linker2, "19852015"))

```


Printing out the dataframe 

```{r comparison graduates table, include=TRUE}
res_combined_graduates <- reduce(res, rbind)

res_combined_graduates %>%
  mutate(across(where(is.numeric), \(x) round(x, digits = 2))) %>%  # this is an updated form of mutate_if()
  kable(format = "latex", digits = 2, booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "scale_down"))

```


Add final number of links to res_combined_graduates

```{r}

res_combined_graduates <- res_combined_graduates |>
  left_join(graduate_counts$links |>
              rename(n_links_final = nb),
            by = c("field" = "fieldname0_mag"))

```



Calculate weighted average of linking stats across fields
 
```{r}
across_fields_graduates <- res_combined_graduates |> 
  left_join(graduate_counts$total |>
              rename(n_graduates = nb),
            by = c("field" = "fieldname0_mag"))

across_fields_graduates <- across_fields_graduates |>
  mutate(share = n_graduates / sum(n_graduates)) |>
  summarise(
    across(all_of(c("same", "only1", "only2", "diff")),
           ~weighted.mean(.x, w = share)),
    across(all_of(c("n_links_final")),
           ~sum(.x))
  ) |>
  mutate(field = "Total")

```


### Make table 

```{r}
df_table_graduates <-  res_combined_graduates %>%  
  select(field, same, only1, only2, diff, n_links_final) %>%
  bind_rows(across_fields_graduates) |> 
  mutate(
    field = stringr::str_to_title((field)),
    across(where(is.numeric), \(x) round(x, digits = 2)) # this is an updated form of mutate_if()
    ) 

```



```{r}

table_caption <- "Linking the graduates"
table_label <- "compare_linking_graduates"
table_columns <- c("Field", "Same entity", "Only by 1", "Only by 2", "Different entity", "Number of links")
footnote <- c(
  "The table summarises the links found from from ProQuest graduates to MAG authors.",
  # the compares the identified links from ProQuest to MAG across training sets by two different labellers.",
  # "The unit of observation is the graduate in ProQuest.",
  "Graduates are defined as the authors of the dissertations in ProQuest.",
  "First, the columns headed by ``Fraction of links found'' compare the identified links across two different labellers as described in the text.",
  "The columns show the fraction of links found for two training sets constructed by two different labellers.",
  "``Same entity'' are graduates for which the models trained on the different training sets find the same MAG identifier.",
  "``Only by 1'' and ``Only by 2'' are graduates for which only the model trained on either of the training sets found a link to MAG at all.",
  "``Different entity'' are graduates for which both models find links to MAG, but to different identifiers.",
  "Second, the last column reports the total number of links found for each field, after all postprocessing (see text for details).",
  "Third, the last row reports the total across fields. The fractions are weighted by the number of graduates in the respective fields in ProQuest 1990--2015."
)
footnote <- paste0(footnote, collapse = " ")

table_graduates <- df_table_graduates |>   
  kable(format = "latex", 
        digits = 2, 
        booktabs = TRUE, 
        caption = table_caption, 
        label = table_label,
        col.names = table_columns
        )  |>
  kableExtra::row_spec(row = length(fields_to_process), hline_after = TRUE) |>
  kableExtra::add_header_above(
    header = c(" " = 1, "Fraction of links found" = 4, " " = 1),
    line = FALSE
  ) |>
  kableExtra::kable_styling(font_size = 9) |>
  kableExtra::footnote(
    general = footnote, 
    footnote_as_chunk = TRUE, 
    threeparttable = TRUE,
    fixed_small_size = TRUE
    ) 

```




```{r}
output_path <- "../../../../output/tables/"
filename <- paste0(output_path, "compare_linking_graduates.tex")
save_kable(table_graduates, file = filename)

```



# Comparison for Advisors


- Compare Christoph and Flavio with Protocol and cleaned institutions


```{r comparison setup advisors, include=TRUE}

fields_to_process <-c( #"art", <- dropped
  "biology", 
  "business",
  "chemistry",
  "computer science" ,
  "economics",
  "engineering",
  "environmental science", 
  "geography", 
  "geology" ,
  "history",
  "materials science",
  "mathematics",
  #"medicine",   <- dropped
  "philosophy",
  "physics",
  "political science",
  "psychology" ,
  "sociology")

linker1 = "flavio_with_protocol_cleaninst" # flavio_with_protocol_cleaninst
linker2 = "christoph_with_protocol_cleaninst" #christoph_degree0_with_protocol_updated
```

```{r comparison advisors, include=FALSE}

# Loop through the fields
res <- lapply(fields_to_process, function(x) compare(x,  "advisors", linker1, linker2 , "19902015"))

```


Printing out the dataframe 


```{r comparison advisors table, include=TRUE}
res_combined_advisors <- reduce(res, rbind)

res_combined_advisors %>%
  mutate(across(where(is.numeric), \(x) round(x, digits = 2))) %>%  # this is an updated form of mutate_if()
  kable(format = "latex", digits = 2, booktabs = TRUE) %>%
  kable_styling(latex_options = c("striped", "scale_down"))

# res_combined_advisors %>%
#   mutate(across(where(is.numeric), \(x) round(x, digits = 2))) %>% View()

```


#### Make table 


Add total final links 

```{r}
res_combined_advisors <- res_combined_advisors |> 
  left_join(advisor_link_counts |>
              rename(n_links_final = nb),
            by = c("field" = "fieldname0_mag"))
```

Summarise across fields

```{r}
across_fields_advisors <- res_combined_advisors |> 
  left_join(graduate_counts$total |>
              rename(n_graduates = nb),
            by = c("field" = "fieldname0_mag"))

across_fields_advisors <- across_fields_advisors |>
  mutate(share = n_graduates / sum(n_graduates)) |>
  summarise(
    across(all_of(c("same", "only1", "only2", "diff")),
           ~weighted.mean(.x, w = share)),
    across(all_of(c("n_links_final")),
           ~sum(.x))
  ) |>
  mutate(field = "Total")
```




```{r}
df_table_advisors <-  res_combined_advisors %>%  
  select(field, same, only1, only2, diff, n_links_final) %>%
  bind_rows(across_fields_advisors) |>
  mutate(
    field = stringr::str_to_title((field)),
    across(where(is.numeric), \(x) round(x, digits = 2)) # this is an updated form of mutate_if()
    ) 
```

```{r}

table_caption <- "Linking the advisors"
table_label <- "compare_linking_advisors"
table_columns <- c("Field", "Same entity", "Only by 1", "Only by 2", "Different entity", "Number of links")
footnote <- c(
  "The table summarises the links found from from ProQuest advisors to MAG authors.",
  "An advisor is one relationship id as described in the text.",
  "First, the columns headed by ``Fraction of links found'' compare the identified links across two different labellers as described in the text.",
  "The columns show the fraction of links found for two training sets constructed by two different labellers.",
  "``Same entity'' are advisors for which the models trained on the different training sets find the same MAG identifier.",
  "``Only by 1'' and ``Only by 2'' are advisors for which only the model trained on either of the training sets found a link to MAG at all.",
  "``Different entity'' are advisors for which both models find links to MAG, but to different identifiers.",
  "Second, the last column reports the total number of links found for each field, after all postprocessing (see text for details).",
  "Third, the last row reports the total across fields. The fractions are weighted by the number of graduates in the respective fields in ProQuest 1990--2015."
)
footnote <- paste0(footnote, collapse = " ")

table_graduates <- df_table_advisors |>   
  kable(format = "latex", 
        digits = 2, 
        booktabs = TRUE, 
        caption = table_caption, 
        label = table_label,
        col.names = table_columns
        )  |>
  kableExtra::row_spec(row = length(fields_to_process), hline_after = TRUE) |>
  kableExtra::add_header_above(
    header = c(" " = 1, "Fraction of links found" = 4, " " = 1),
    line = FALSE
  ) |>
  kableExtra::kable_styling(font_size = 9) |>
  kableExtra::footnote(
    general = footnote, 
    footnote_as_chunk = TRUE, 
    threeparttable = TRUE,
    fixed_small_size = TRUE
    ) 

```


```{r}
output_path <- "../../../../output/tables/"
filename <- paste0(output_path, "compare_linking_advisors.tex")
save_kable(table_graduates, file = filename)
```



# Compare names manually

```{r}

field="chemistry"
linktype="advisors"
years="19902015"

links <- compare(field, linktype, linker1, linker2, years, inspect=TRUE)



```

```{r, include = FALSE}
DBI::dbDisconnect(con)
```
