---
title: "Performance of linking researchers to theses"
author: "Flavio & Christoph & Mona"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
---

This script makes some plots of the advisor links and saves the most plausible links to a table in the database. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

```{r, echo = FALSE, include = FALSE}
packages <- c("tidyverse", "broom", "dbplyr", "RSQLite", "ggplot2", "stringdist")

lapply(packages, library, character.only = TRUE)

datapath <- "/mnt/ssd/"
db_file  <- paste0(datapath, "AcademicGraph/AcademicGraph.sqlite")
select_fields <- c("biology",
                    "business",
                    "chemistry",
                    "computer science" ,
                    "economics",
                    "engineering",
                    "environmental science",
                    "geography",
                    "geology" ,
                    "history",
                    "materials science",
                    "mathematics",
                    "physics",
                    "political science",
                    "philosophy",
                    "psychology" ,
                    "sociology") # art, medicine linked but not included yet 


# ## db connection
con <- DBI::dbConnect(RSQLite::SQLite(), db_file)
cat("The database connection is: \n")
src_dbi(con)
```



```{r}
# parameters for selecting links 
min_score_advisors <- 0.7 # minimum score from dedupe 

```

Linking Mona vs Christoph
```{r}
# Function to process the data for a specific field
field <- c("sociology")
#"biology", "business", "chemistry", "computer science", "economics", "engineering", "environmental science", "geography", "geology", "history", "materials science", "mathematics", "philosophy", "physics", "political science", "psychology", "sociology"

  
  # Read the data for the specified field
  links_advisors_mona <- read.csv(paste0("/mnt/ssd/linked_ids_temp/links_advisors_", field, "_mona_degree0_19902015.csv")) %>%
    rename(relationship_id = grantid_authorposition)

  links_advisors_christoph <- read.csv(paste0("/mnt/ssd/linked_ids_temp/links_advisors_", field, "_christoph_degree0_19902015.csv"))

  linked_advisors_mona <- collect(links_advisors_mona)
  linked_advisors_christoph <- collect(links_advisors_christoph)

  # Perform the left join
  linked_advisors <- linked_advisors_mona %>%
    left_join(linked_advisors_christoph, by = c("AuthorId", "relationship_id")) %>%
    mutate(difference = ifelse(!is.na(link_score.x) & !is.na(link_score.y), link_score.x - link_score.y, NA))
  
  linked_advisors %>%
    ggplot(aes(x = difference)) +
    geom_histogram(bins = 100, aes(y = after_stat(density))) 
```

business: okay?
econ: almost same
engineering, geography, materials science, physics, political science, sociology problem?




## Overview

```{r, include = FALSE}

linking_info <- tbl(con,
                    sql("select iteration_id, field
                          from (
                              select *, max(iteration_id) OVER(partition by field) as max_iter
                              from linking_info_advisors
                              where recall = 0.9 
                                  and testing = 0
                                  and institution = 'True'
                                  and fieldofstudy_cat = 'False'
                                  and fieldofstudy_str = 'False' 
                                  and keywords = 'False' 
                                  and mergemode = 'm:1'
                          )
                          where iteration_id = max_iter "))
linked_advisors <- tbl(con, sql("select * 
                                from linked_ids_advisors
                                ")) %>%
  inner_join(linking_info %>% 
               select(iteration_id),
             by = "iteration_id")



theses <- tbl(con, sql("
                       select * from pq_advisors 
                       inner join (
                         select goid, degree_year, uni_name 
                         from pq_authors
                         inner join (
                          select university_id, normalizedname as uni_name 
                          from pq_unis
                          where location like '%United States%'
                         ) using(university_id)
                       ) using(goid)
                       inner join (
                        select goid, normalizedname as fieldname0_mag
                        from pq_fields_mag
                        inner join(
                          select fieldofstudyid, normalizedname
                          from fieldsofstudy
                        ) on (mag_field0 = fieldofstudyid)
                        where position = 0
                       ) using(goid)
                       ")
              )

pq_fields_mag <- tbl(con,
                     sql("select goid, normalizedname as field 
                          from pq_fields_mag
                          inner join (
                            select fieldofstudyid, normalizedname
                            from FieldsOfStudy
                            ) ON (mag_field0 = FieldOfStudyId)
                         where position = 0"))

```


```{r}
linked_advisors <- collect(linked_advisors)
theses <- collect(theses)
linking_info <- collect(linking_info)
pq_fields_mag <- collect(pq_fields_mag)

```


## Linking scores 

- conditioning on link score > 0.7 is fine 
- 

```{r}

linked_advisors %>%
  left_join(linking_info, by = "iteration_id") %>%
  filter(link_score>=0.7) %>%
  ggplot(aes(x = link_score)) +
  geom_histogram(bins = 100, aes( y = after_stat(density))) +
  facet_wrap(~field)

```

## Link performance by graduation year

- fraction of listed advisors where the link_score is above the treshold
- the mean link score for advisors where dedupe finds a link (link_score is not NA)
- NOTE: the field here is assigned based on the first reported in the dissertation, and the crosswalked to the MAG field
- in the figure above, we used the field from iteration_id, but this only works for advisors that dedupe suggests to be a link 

```{r}

keep_fields <- select_fields
# c("biology", "chemistry", "computer science", 
#                 "economics", "engineering", "environmental science", 
#                 "geography", "geology", "mathetmatics", "physics",
#                 "political science", "psychology", "sociology")

score_by_year <- theses %>% 
  filter(degree_year >= 1990) %>%
  left_join(linked_advisors,
            by = "relationship_id") %>%
  left_join(pq_fields_mag, by = "goid") %>%
  filter(field %in% keep_fields)

# %>%
#   left_join(linking_info,
#             by = "iteration_id")

score_by_year %>%
  mutate(link_score_adj = ifelse(is.na(link_score), -1, link_score)) %>%
  group_by(degree_year, field) %>%
  summarise(mean_score = mean(link_score, na.rm = TRUE),
            #p50_score = quantile(link_score, probs = 0.5),
            share_linked = mean(link_score_adj > min_score_advisors),
            .groups = "drop") %>%
  pivot_longer(cols = all_of(c("mean_score", "share_linked")),
               names_to = "stat") %>%
  ggplot(aes(x = degree_year, y = value)) + 
  geom_line(aes(linetype = stat)) + 
  facet_wrap(~field) + 
  theme(legend.position = "bottom")

```




```{r, include = FALSE}
DBI::dbDisconnect(con)
```
