---
title: "Comparing links in chemistry"
author: "Flavio & Christoph"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
---

This document compares the final links we obtain for chemistry.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)

packages <- c("tidyverse", "broom", "dbplyr", "RSQLite", "ggplot2", "lubridate", "magutils")

lapply(packages, library, character.only = TRUE)

datapath <- "/mnt/ssd/"
db_file  <- paste0(datapath, "AcademicGraph/AcademicGraph.sqlite")

```

```{r}
select_fields <- c("biology",
                   "business",
                   "chemistry",
                   "computer science" ,
                   "economics",
                   "engineering",
                   "environmental science",
                   "geography",
                   "geology" ,
                   "history",
                   "materials science",
                   "mathematics",
                   "medicine",
                   "art",
                   "philosophy",
                   "physics",
                   "political science",
                   "psychology" ,
                   "sociology") 

```


```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), db_file)
cat("The database connection is: \n")
src_dbi(con)
```

```{r}
field_names_id <- tbl(con, sql(paste0(
    "SELECT FieldOfStudyId, NormalizedName
    FROM FieldsOfStudy
    WHERE Level = 0 
        AND NormalizedName IN (", 
        paste0(paste0("'", select_fields, "'"), collapse = ", "),
        ")"
)))
field_names_id <- collect(field_names_id)


query_mag <- paste0(
 "SELECT AuthorId
        , year
        , fieldofstudy
        , mag_field0 
    FROM (
        SELECT a.AuthorId
            , a.YearFirstPub AS year
            , e.NormalizedName AS fieldofstudy
            , e.ParentFieldOfStudyId as mag_field0
        FROM author_sample AS A 
        INNER JOIN (
            SELECT AuthorId, NormalizedName, ParentFieldOfStudyId
            FROM author_fields c
            INNER JOIN (
                SELECT FieldOfStudyId, NormalizedName
                FROM FieldsOfStudy
            ) AS d USING(FieldOfStudyId)
            INNER JOIN (
                SELECT ParentFieldOfStudyId
                    , ChildFieldOfStudyId
                    , ParentFieldOfStudyId 
                FROM crosswalk_fields
                WHERE ParentLevel = 0
                    AND ParentFieldOfStudyId IN (",
                         paste0(field_names_id$FieldOfStudyId, collapse = ", "),
                    ")
            ) AS e ON (e.ChildFieldOfStudyId = c.FieldOfStudyId)
            WHERE FieldClass = 'first'
        ) AS e USING(AuthorId)
    )
WHERE year >= 1980 and year <= 2022 
")
```




```{r}

linked_ids <- tbl(con, "current_links") 
linking_info <- tbl(con, "linking_info") %>%
    filter(mergemode == "1:1" & fieldofstudy_str == "False")
pq_authors <- get_proquest(conn = con, from = "graduates", start_year = 1990, end_year = 2015)
mag_authors <- tbl(con, sql(query_mag))


```


```{r, include = FALSE}
mag_authors <- collect(mag_authors)
linked_ids <- collect(linked_ids)
linking_info <- collect(linking_info)
pq_authors <- collect(pq_authors)
```


Combine data pq and MAG

```{r}

d_linked <- linked_ids |> 
  select(AuthorId, goid) |> 
  left_join(pq_authors |>
              select(-gender),
            by = "goid") |> 
  left_join(mag_authors |>
              select(AuthorId, year_firstpub = year),
            by = "AuthorId")

```


# Chemistry: first affiliation of MAG authors should be the graduating institution. [paper](https://direct.mit.edu/rest/article-abstract/95/2/698/58091/Chinese-Graduate-Students-and-U-S-Scientific?redirectedFrom=fulltext)


```{r}
d_main <- d_linked |> 
  filter(fieldname0_mag == "chemistry") |> 
  mutate(grp = case_when( # some people publish already way before the PhD
    year_firstpub > degree_year ~ "first pub after PhD",
    year_firstpub < degree_year - 6 ~ "first pub before PhD",
    TRUE ~ "first pub during PhD"
  )) |>
  select(AuthorId, goid, degree_year, grp)
  
  
```


```{r}

d_main |> 
  group_by(grp, degree_year) |> 
  summarise(nb = n()) |>
  ungroup() |>
  group_by(degree_year) |>
  mutate(total = sum(nb)) |>
  ggplot(aes(x = degree_year, y = nb/total)) + 
  geom_line(aes(linetype = grp)) +
  theme(legend.position = "bottom")

```

Gaule/Piacentini had 21154 graduates from 1999 to 2008; we have 

```{r}

d_main |> 
  filter(degree_year >= 1999 & degree_year <= 2008) |> 
  summarise(n())

```
- they had chemists and chemical engineers; we miss the engineers in this sample.


```{r}

query_authors <- unique(d_main$AuthorId)
query_authors <- paste0(query_authors, collapse = ", ")
q_authors_affil <- paste0(
  "SELECT AuthorId, AffiliationId, Year
  FROM AuthorAffiliation
  INNER JOIN (
    SELECT AuthorId, YearFirstPub
    FROM author_sample
  ) USING(AuthorId)
  WHERE AuthorId IN (", query_authors, ")
  AND Year <= YearFirstPub + 20"
)

authors_affil <- tbl(con, sql(q_authors_affil)) |>
  collect() 

authors_first_affil <- authors_affil |>
  group_by(AuthorId) |>
  filter(Year == min(Year)) |>
  filter(!duplicated(AuthorId)) |>
  ungroup()



```


```{r}
links_to_cng <- tbl(con, "links_to_cng") |>
  collect()
```


### Place of first publication

```{r}
place_first_pub <- d_main |>
  left_join(pq_authors |>
              select(goid, university_id),
            by = "goid") |>
  inner_join(links_to_cng |>
              filter(from_dataset == "pq") |>
              select(from_id, unitid_graduate = unitid),
            by = c("university_id" = "from_id")) |>
  left_join(authors_first_affil |>
              select(AuthorId, AffiliationId),
            by = "AuthorId") |>
  inner_join(links_to_cng |>
              filter(from_dataset == "mag") |>
              select(from_id, unitid_author = unitid),
            by = c("AffiliationId" = "from_id"))
```

```{r}
place_first_pub |>
  mutate(same_institution = ifelse(unitid_graduate == unitid_author, 1, 0)) |>
  group_by(degree_year) |>
  summarise(same_institution = mean(same_institution, na.rm = T),
            .groups = "drop") |>
  ggplot(aes(x = degree_year, y = same_institution)) +
  geom_line() +
  geom_point()

```

### If publishing during PhD, does so at least once at the PhD university?

```{r}
publish_during_phd <- authors_affil |>
  left_join(d_main |>
              select(-grp),
            by = c("AuthorId")) |>
  filter(Year <= degree_year & Year >= degree_year - 6) |>
  inner_join(links_to_cng |>
              filter(from_dataset == "mag") |>
              select(from_id, unitid_author = unitid),
            by = c("AffiliationId" = "from_id")) |>
  left_join(pq_authors |>
              select(goid, university_id),
            by = "goid") |>
  inner_join(links_to_cng |>
              filter(from_dataset == "pq") |>
              select(from_id, unitid_graduate = unitid),
            by = c("university_id" = "from_id")) |>
  select(AuthorId, Year, degree_year, unitid_author, unitid_graduate, university_id) |>
  mutate(same_institution = ifelse(unitid_author == unitid_graduate, 1, 0),
         same_institution = ifelse(is.na(same_institution), 0, same_institution))

```


Fraction of students not publishing during PhD:
```{r}
1 - n_distinct(publish_during_phd$AuthorId) / n_distinct(d_main$AuthorId)
```

```{r}
# group by student: at least one pub with the PhD university?
publish_during_phd <- publish_during_phd |>
  group_by(AuthorId) |>
  filter(same_institution == max(same_institution)) |>
  filter(!duplicated(AuthorId))

publish_during_phd |>
  group_by(degree_year) |>
  summarise(same_institution = mean(same_institution, na.rm = T),
            .groups = "drop") |>
  ggplot(aes(x = degree_year, y = same_institution)) +
  geom_line() +
  geom_point()
  

```

```{r}

summary(publish_during_phd)

```

Notes

- some may publish after phd with the phd affiliation -- not captured here 
- misses 
  - research institutes that are not in Carnegie, ie scripps research institute
  - chemical engineers
- all in all, this is a lower bound on the precision in the sample of chemists publishing during their PhD
- the lower bound on precision for the sample of chemists can be calculated as follows
  - 18% publish after PhD; assume they are all false positives 
  - of the remaining 82%, 96% publish at their graduating university
  - thus, our precision is at least 0.82 * 0.96 = 0.78
- this calculation is more difficult in fields where graduates publish more often after graduating



```{r, include = FALSE}
DBI::dbDisconnect(con)
```

