---
title: "Performance of linking researchers to theses"
author: "Flavio & Christoph"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
---

This script makes some plots of the advisor links and saves the most plausible links to a table in the database. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

```{r, echo = FALSE, include = FALSE}
packages <- c("tidyverse", "broom", "dbplyr", "RSQLite", "ggplot2", "stringdist")

lapply(packages, library, character.only = TRUE)

datapath <- "/mnt/ssd/"
db_file  <- paste0(datapath, "AcademicGraph/AcademicGraph.sqlite")
select_fields <- c("physics", "biology", "chemistry", "sociology",
                    "economics", "political science", "psychology", 
                    "mathematics", "geography", "geology", "engineering",
                    "computer science", "environmental science") # fields currently matched 


# ## db connection
con <- DBI::dbConnect(RSQLite::SQLite(), db_file)
cat("The database connection is: \n")
src_dbi(con)
```



```{r}
# parameters for selecting links 
min_score_advisors <- 0.7 # minimum score from dedupe 

```


## Overview

```{r, include = FALSE}

linking_info <- tbl(con,
                    sql("select iteration_id, field
                          from (
                              select *, max(iteration_id) OVER(partition by field) as max_iter
                              from linking_info_advisors
                              where recall = 0.9 
                                  and testing = 0
                                  and institution = 'True'
                                  and fieldofstudy_cat = 'False'
                                  and fieldofstudy_str = 'False' 
                                  and keywords = 'False' 
                                  and mergemode = 'm:1'
                          )
                          where iteration_id = max_iter "))
linked_advisors <- tbl(con, sql("select * 
                                from linked_ids_advisors
                                ")) %>%
  inner_join(linking_info %>% 
               select(iteration_id),
             by = "iteration_id")



theses <- tbl(con, sql("
                       select * from pq_advisors 
                       inner join (
                         select goid, degree_year, uni_name 
                         from pq_authors
                         inner join (
                          select university_id, normalizedname as uni_name 
                          from pq_unis
                          where location like '%United States%'
                         ) using(university_id)
                       ) using(goid)
                       inner join (
                        select goid, normalizedname as fieldname0_mag
                        from pq_fields_mag
                        inner join(
                          select fieldofstudyid, normalizedname
                          from fieldsofstudy
                        ) on (mag_field0 = fieldofstudyid)
                        where position = 0
                       ) using(goid)
                       ")
              )

pq_fields_mag <- tbl(con,
                     sql("select goid, normalizedname as field 
                          from pq_fields_mag
                          inner join (
                            select fieldofstudyid, normalizedname
                            from FieldsOfStudy
                            ) ON (mag_field0 = FieldOfStudyId)
                         where position = 0"))

```


```{r}
linked_advisors <- collect(linked_advisors)
theses <- collect(theses)
linking_info <- collect(linking_info)
pq_fields_mag <- collect(pq_fields_mag)

```


## Linking scores 

- conditioning on link score > 0.7 is fine 

```{r}

linked_advisors %>%
  left_join(linking_info, by = "iteration_id") %>%
  ggplot(aes(x = link_score)) +
  geom_histogram(bins = 100, aes( y = ..density..)) +
  facet_wrap(~field)

```

## Link performance by graduation year

- fraction of listed advisors where the link_score is above the treshold
- the mean link score for advisors where dedupe finds a link (link_score is not NA)
- NOTE: the field here is assigned based on the first reported in the dissertation, and the crosswalked to the MAG field
  - in the figure above, we used the field from iteration_id, but this only works for advisors that dedupe suggests to be a link 

```{r}

keep_fields <- c("biology", "chemistry", "computer science", 
                 "economics", "engineering", "environmental science", 
                 "geography", "geology", "mathetmatics", "physics",
                 "political science", "psychology", "sociology")

score_by_year <- theses %>% 
  filter(degree_year >= 1985) %>%
  left_join(linked_advisors,
            by = "relationship_id") %>%
  left_join(pq_fields_mag, by = "goid") %>%
  filter(field %in% keep_fields)

# %>%
#   left_join(linking_info,
#             by = "iteration_id")

score_by_year %>%
  mutate(link_score_adj = ifelse(is.na(link_score), -1, link_score)) %>%
  group_by(degree_year, field) %>%
  summarise(mean_score = mean(link_score, na.rm = TRUE),
            #p50_score = quantile(link_score, probs = 0.5),
            share_linked = mean(link_score_adj > min_score_advisors),
            .groups = "drop") %>%
  pivot_longer(cols = all_of(c("mean_score", "share_linked")),
               names_to = "stat") %>%
  ggplot(aes(x = degree_year, y = value)) + 
  geom_line(aes(linetype = stat)) + 
  facet_wrap(~field) + 
  theme(legend.position = "bottom")

```


```{r}

score_by_year %>%
  filter(field == "biology") %>% 
  filter(is.na(link_score)) %>% 
  filter(degree_year < 2000) %>% 
  group_by(firstname, lastname, uni_name, degree_year) %>%
  summarise(nb = n(),
            .groups = "drop") %>%
  arrange(desc(nb)) %>%
  head(10)

# score_by_year %>% filter(lastname == "dasgupta" & firstname == "asim" & !is.na(iteration_id)) # never linked 
# score_by_year %>% filter(lastname == "freeling" & firstname == "michael") # never linked 

# scale this up? check all the main fields of the authors with such names? -- tedious 
```

### Notes 


- Reasons for why advisor not linked
  - they are not sampled for linking either in the mag or proquest data
    - most plausibly because they are assigned to different fields 
  - institution names do not overlap 
  - dedupe does not find a link even though it should
    - but how can it explain the time trend?
- Comparing fields in MAG and ProQuest dissertations
  - General
    - not linking an advisor in biology does not mean do not link them in chemistry if the thesis is also classified in chemistry
    - in the data above, this happens if biology is listed at position 0
  - Biology 
    <!-- - main field chemistry: gerlt, cooperman, eveleigh (two of them with long careers, but both in chemistry), tsai -->
    - main field biology: dasgupta, freeling
    - at least one of the dissertations of freeling are sampled for the linking 
  - Sociology
    - different main field: ishisaka, coulton (medicine), howell (geography), mindel (psychology)
    - not in MAG, but findable on google: khleif, gullerud
    - not in MAG, not findable on google: liff
- Next steps
  - widen the sampled field in MAG 
  - re-train and re-check 


### Update, 4/11/22
- c brent theurer (authorid 2005171991), c channa reddy (authorid 2263585539): not in advisor sample b/c of the constraint on the length of the first name. can we fix this?
- john a gerlt: similar; is in mag with j a or ja gerlt
- daniel rittschof: his author id is 2242600877 and his name there is dan rittschof
- mingdaw tsai (authorid 2159629249), naba k gupta (authorid 2298396241), paul f cook (authorid 2107503814), eric n olson (2029316736) are all in the advisor sample from mag
  - do their affiliation-years exactly match in MAG? all except olson, who has a special affiliation    
  - what are the possibilities then?
    - algorithm
      - blocking: predicates are `(SimplePredicate: (commonSixGram, lastname), SimplePredicate: (commonFourGram, middlename))`. In contrast, for the settings file 1985-2022, the predicates are `(SimplePredicate: (suffixArray, lastname),)`. For chemistry, they are `(SimplePredicate: (sameSevenCharStartPredicate, lastname),)`
        - why do we use another training data set again here?
      - logistic regression
      - they are false negatives. compared to chemistry, it seems weird to have such a high false negative rate (this assumes that all the advisors we do currently not find are actually in the data)
    - training
    - the features are wrong. -- No, they are correct. 
    - the comparator is wrong. but then why does it work for other fields?
  - any reason should ideally also explain why it works for chemistry but not for biology
- How can we fix the algorithm?
  - we could relax the feature "same firstname"/"same lastname", particularly for advisors where the affiliation is a good and precise feature



Here is some python code to look at the learned settings, based on

- https://github.com/dedupeio/rlr/blob/master/rlr/lr.py (new dedupe does not use this anymore I think)
- https://github.com/dedupeio/dedupe/blob/5742efc7fc696c06d3327e038541532e584551a8/dedupe/api.py
- Note: The predicates are similar for all three fields I looked at. I do not know how the weights correspond to the logit regression coefficients

```python

sf_biology = "/mnt/ssd/DedupeFiles/advisors/settings_biology_1985_2022_institutionTrue_fieldofstudy_catFalse_fieldofstudy_strFalse_keywordsFalsechristoph_baseline_update"
sf_chemistry = "/mnt/ssd/DedupeFiles/advisors/settings_chemistry_1985_2022_institutionTrue_fieldofstudy_catFalse_fieldofstudy_strFalse_keywordsFalsechristoph_baseline_update"
sf_cs = "/mnt/ssd/DedupeFiles/advisors/settings_computer_science_1985_2022_institutionTrue_fieldofstudy_catFalse_fieldofstudy_strFalse_keywordsFalsechristoph_baseline_update"

with open(sf_biology, "rb") as sf:
    linker_biology = dedupe.StaticRecordLink(sf)

with open(sf_chemistry, "rb") as sf:
    linker_chemistry = dedupe.StaticRecordLink(sf)


with open(sf_cs, "rb") as sf:
    linker_cs = dedupe.StaticRecordLink(sf)

linker_biology.predicates
linker_chemistry.predicates
linker_cs.predicates

linker_biology.classifier.weights
linker_chemistry.classifier.weights
linker_cs.classifier.weights

```


sf_biology = "/mnt/ssd/DedupeFiles/advisors/settings_biology_1985_2022_institutionTrue_fieldofstudy_catFalse_fieldofstudy_strFalse_keywordsFalsechristoph_degree0"


## Number of linked advisors 

- not sure this is still relevant?

```{r}
score_by_year %>%
  filter(!is.na(link_score)
          & field %in% keep_fields) %>%
  group_by(degree_year, field) %>%
  summarise(n_advisors = n(),
            .groups = "drop") %>%
  ggplot(aes(x = degree_year, y = n_advisors)) + 
  geom_line() + 
  facet_wrap(~field)

```

old comments

- for instance, a student of michael j lambert (authorid 2120159045; relationship id 303670971_0 in proquest) from pre-1990 is link score of 0.02, but should be a clear link


## Compare number of links across iterations within fields 

```{r}
fields_iter_compare <- c("economics", "chemistry")
min_score <- 0.8

keep_iter_ids <- tbl(con, "linking_info_advisors") %>%
  filter(field %in% fields_iter_compare) %>%
  filter(testing == 0) %>%
  collect() %>%
  group_by(field, train_name) %>%
  arrange(iteration_id) %>%
  mutate(nb = n(),
         id = row_number()) %>%
  ungroup() %>%
  filter(id == nb) %>%
  select(iteration_id, field, train_name)

linked_ids_to_compare <- tbl(con, "linked_ids_advisors") %>%
  inner_join(
    tbl(con, "linking_info_advisors") %>%
      filter(field %in% fields_iter_compare),
    by = "iteration_id"
  ) %>%
  inner_join(
    tbl(con, "pq_advisors") %>%
      select(relationship_id, goid),
    by = "relationship_id"
  ) %>%
  inner_join(
    tbl(con, "pq_authors") %>%
      select(goid, degree_year),
    by = "goid"
  ) %>%
  collect() %>%
  filter(iteration_id %in% keep_iter_ids$iteration_id) 

```


Number of graduates with at least 1 advisor

```{r}

d_sum <- linked_ids_to_compare %>%
  filter(link_score >= min_score) %>%
  group_by(train_name, field, degree_year) %>%
  summarise(n_advisors = n(),
            n_graduates = n_distinct(goid),
            .groups = "drop") %>%
  pivot_longer(cols = starts_with("n_"), names_to = "variable")

plotvars <- c("n_graduates")

map(.x = plotvars,
    .f = ~d_sum %>%
      filter(variable == .x) %>%
      ggplot(aes(x = degree_year, y = value)) +
      geom_line(aes(linetype = train_name)) + 
      facet_wrap(~field) + 
      theme(legend.position = "bottom") + 
      labs(title = paste0("Count: ", .x))
    )


```

## Fraction of theses with at least 1 supervisor linked to MAG

```{r}
s_thesis_advisor_link <- theses %>%
  filter(degree_year %in% 1990:2015) %>%
  left_join(linked_advisors %>%
              filter(link_score > min_score_advisors) %>%
              select(relationship_id) %>%
              mutate(linked = 1),
            by = "relationship_id") %>%
  mutate(linked = ifelse(is.na(linked), 0, linked)) %>%
  group_by(goid) %>%
  mutate(any_link = max(linked)) %>%
  ungroup() %>%
  filter(!duplicated(goid)) %>%
  group_by(degree_year, any_link, fieldname0_mag) %>%
  summarise(n_theses = n(), 
            .groups = "drop") %>%
  group_by(degree_year, fieldname0_mag) %>%
  mutate(s = n_theses / sum(n_theses)) %>%
  ungroup() %>%
  filter(any_link == 1)
```


```{r}

s_thesis_advisor_link %>% 
  ggplot(aes(x = degree_year, y = s)) +
  geom_line() +
  facet_wrap(~fieldname0_mag) + 
  scale_x_continuous(breaks = c(1995, 2005, 2015)) + 
  geom_hline(yintercept = 1, color = "grey55", linetype = "dashed") +
  labs(y = "Share", title = "Share of US theses with >=1 supervisor linked to MAG",
       subtitle = "Where >= 1 supervisor reported; by first indicated field0 of thesis.")

```

### Notes

- Idea: since supervisors tend to be established researchers and publish regularly, we should find a large fraction of supervisors reported in ProQuest in the MAG data.

- The split by field is not exact because the link may have been found using a different reported field0.

- The close to 100% is reassuring of the MAG data quality on affiliations in these fields.

- Fields of concern: physics, sociology, poli science, biology (the level, the break and the trend).



## Note: the "usable" links are saved to the db in src/dataprep/main/link/prep_linked_data.py



# Check whether the entities exist at all in the underlying data from MAG 

Copy-paste the sql query for now


```{r}
fieldofstudy_id <- 86803240
query_mag <- paste0(
  "SELECT f.AuthorId
        , f.year
        , f.YearLastPub
        , f.firstname
        , f.lastname
        , CASE TRIM(SUBSTR(f.middle_lastname, 1, f.l_fullname - f.l_firstname - f.l_lastname - 1)) 
            WHEN 
                '' THEN NULL 
                ELSE TRIM(SUBSTR(f.middle_lastname, 1, f.l_fullname - f.l_firstname - f.l_lastname - 1)) 
            END as middlename 
        , f.fieldofstudy
        , g.keywords
        , g.coauthors
        , g.institution
        , g.main_us_institutions_year
    FROM (
        SELECT a.AuthorId
            , a.YearFirstPub AS year
            , a.YearLastPub 
            , a.FirstName AS firstname
            , REPLACE(b.NormalizedName, RTRIM(b.NormalizedName, REPLACE(b.NormalizedName, ' ', '')), '') AS lastname 
            , TRIM(SUBSTR(b.NormalizedName, length(a.FirstName) + 1)) AS middle_lastname 
            , length(b.NormalizedName) as l_fullname 
            , length(a.FirstName) as l_firstname
            , length(REPLACE(b.NormalizedName, RTRIM(b.NormalizedName, REPLACE(b.NormalizedName, ' ', '')), '')) as l_lastname
            , e.NormalizedName AS fieldofstudy
        FROM author_sample AS a
        INNER JOIN (
            SELECT AuthorId, NormalizedName
            FROM Authors
        ) AS b USING(AuthorId)
        INNER JOIN (
            SELECT AuthorId
            FROM author_field0
            WHERE FieldOfStudyId_lvl0 =", fieldofstudy_id, "
                AND Degree <= 0
        ) USING(AuthorId)
        LEFT JOIN (
            SELECT AuthorId, NormalizedName
            FROM author_fields c
            INNER JOIN (
                SELECT FieldOfStudyId, NormalizedName
                FROM FieldsOfStudy
            ) AS d USING(FieldOfStudyId)
            INNER JOIN (
                SELECT ParentFieldOfStudyId, ChildFieldOfStudyId
                FROM crosswalk_fields
                WHERE ParentLevel = 0
                    AND ParentFieldOfStudyId = ", fieldofstudy_id, "
            ) AS e ON (e.ChildFieldOfStudyId = c.FieldOfStudyId)
            WHERE FieldClass = 'first'
        ) AS e USING(AuthorId)
    ) f
    LEFT JOIN (
        SELECT AuthorId
                , main_us_institutions_career as institution
                , coauthors
                , keywords
                , main_us_institutions_year
                , all_us_institutions_year
        FROM author_info_linking
    ) AS g USING(AuthorId)
    WHERE length(firstname) > 1 AND f.YearLastPub >= 1985 AND year <= 2015 + 5 AND institution is not NULL
    AND institution not like '%chinese academy of sciences%'
    "
)

advisor_sample_mag <- tbl(con, sql(query_mag))
advisor_sample_mag <- collect(advisor_sample_mag)

```




```{r}
unlinked_advisors <- score_by_year %>%
  filter(field == "biology"
          & is.na(link_score)) %>% 
  group_by(firstname, lastname, uni_name, degree_year) %>%
  summarise(nb = n(),
            .groups = "drop") 

dk <- unlinked_advisors %>%
  left_join(advisor_sample_mag %>%
              select(AuthorId, year, firstname, lastname, main_us_institutions_year),
            by = c("firstname", "lastname", "degree_year" = "year"))


```


```{r}
head(dk %>% filter(is.na(AuthorId) & degree_year <= 2015) %>% arrange(desc(nb)), 10)
```


Do these entities exist in MAG?
- andrew fire 2012 in stanford. yes. not in advisor sample. authorid 683352831.
- benjamin c start 2002 at iit. yes. has some duplicates. is in advisor sample. authorid 1992276655.
- garry nolan 2010 stanford. yes. not in advisor sample. authorid 1989754750.
- 


```{r}
head(dk %>% filter(is.na(AuthorId) & degree_year <= 2000) %>% arrange(desc(nb)), 10)
```

- dagmar ringe 2000 brandeis. yes. has some duplicates. not in advisor sample. authorid 2171354986
- naba k gupta 1997 at nebraska lincoln. yes. has some duplicates. not in advisor sample. authorid 2298396241

## Why are so many not in the sample? note they are all in author_sample
- not in biology. no, they all have field level 0 biology. 
- none of them has chinese academy of sciences affiliation
- other filtering
  - all of the examples above start their publishing career before 1985. Our query for the linking wrongly filters them out.


```{r, include = FALSE}
DBI::dbDisconnect(con)
```


Here is some sql code that I used for checking the cases of the non-linked biology advisors


```sql

-- 1. they are not in the sample 
-- 2. they are not recognized as links: (a) the model is wrong, (b) the data are wrong (ie, dedupe is confused by the dupes in MAG)
-- the fact that the same entities are not linked even after our improvements suggests perhaps that we are in case 2.b)? but then, dasgupta has no duplicates in MAG

-- from older checks:
-- john a gerlt: actually registered as "j a gerlt" in MAG (authorid 93129757). dedupe links him to authorid 3129783099 in 2000, 2009, 2010, 2014. this is b/c of the duplicate entity
-- asim dasgupta: authorid 2150423063; 

-- authors: 2298396241, 2171354986, 683352831, 1989754750

-- all of them are in biology
select count(distinct authorid)
from author_field0 a
inner join (
    select fieldofstudyid, normalizedname
    from fieldsofstudy
) b on (a.fieldofstudyid_lvl0 = b.fieldofstudyid)
where authorid in (2298396241, 2171354986, 683352831, 1989754750) 
and a.Degree = 0
and normalizedname = "biology"

-- none of them are in chinese academy of sciences
select count(*)
from author_info_linking
where authorid in (2298396241, 2171354986, 683352831, 1989754750)
and main_us_institutions_career not like "chinese academy of sciences"

-- only two have no missing information on the first field. but is this relevant? -> clearly not, as otherwise these two would still be in the advisor sample
SELECT AuthorId, NormalizedName
FROM author_fields c
INNER JOIN (
    SELECT FieldOfStudyId, NormalizedName
    FROM FieldsOfStudy
) AS d USING(FieldOfStudyId)
-- ## Condition on fieldofstudy being in the level 0 id_field
INNER JOIN (
    SELECT ParentFieldOfStudyId, ChildFieldOfStudyId
    FROM crosswalk_fields
    WHERE ParentLevel = 0
        AND ParentFieldOfStudyId IN (86803240)
) AS e ON (e.ChildFieldOfStudyId = c.FieldOfStudyId)
WHERE FieldClass = 'first'
and authorid in (2298396241, 2171354986, 683352831, 1989754750)
limit 10;

-- they all start publishing before 1985
select * 
from author_sample 
where authorid in (2298396241, 2171354986, 683352831, 1989754750)

```

