---
title: "Some exploration of ProQuest data quality"
author: "Flavio & Christoph"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

```{r, echo = FALSE, include = FALSE}
packages <- c("tidyverse", "broom", "dbplyr", "RSQLite", "ggplot2", "data.table")

lapply(packages, library, character.only = TRUE)

datapath <- "/mnt/ssd/"
db_file  <- paste0(datapath, "AcademicGraph/AcademicGraph.sqlite")

# ## db connection
con <- DBI::dbConnect(RSQLite::SQLite(), db_file)
cat("The database connection is: \n")
src_dbi(con)
```


```{r}

q <- "select * from pq_authors 
        left join (
                select university_id, location
                from pq_unis
        ) using (university_id)
        "

authors <- tbl(con, sql(q)) %>%
    collect() %>% # focus only on U.S. institutions
    filter(grepl("United States", location))

advisors <- tbl(con, "pq_advisors") %>%
    collect()

fields <- tbl(con, "pq_fields") %>%
    collect()

q <- "select * from pq_fields_mag a
        inner join (select FieldOfStudyId, NormalizedName as fieldname_mag 
                    from FieldsOfStudy) b on (a.mag_field0 = b.FieldOfStudyId)"
fields_mag <- tbl(con, sql(q)) %>%
    collect()


```


```{r}

path_nces <- paste0(datapath, "NCES_NSF/processed/") 

data_graduates <- fread(paste0(path_nces, "graduate_counts.csv"), data.table = FALSE)
fields_nces <- fread(paste0(path_nces, "fields.csv"), data.table = FALSE)

data_graduates <- data_graduates %>%
  left_join(fields_nces %>%
              select(field_id, fld = shortname),
            by = "field_id") %>%
  select(-field_id)

```


```{r}

# # correspond the fields to the mag_field0
#     ## this does not work b/c correspondence to fields is not very good 

# direct_detailed <- c("economics", "psychology",
#                      "sociology", "history", "chemistry")
# data_graduates <- data_graduates %>%
#     filter(det_fld != "total" 
#             & broad_fld != "total"
#             & se_fld != "total") %>%
#     mutate(mag_field0 = case_when(
#         det_fld %in% direct_detailed ~ det_fld, # detailed fields with exact match
#         det_fld == "political science and government" ~ "political science",
#         det_fld == "computer and information sciences" ~ "computer science",
#         det_fld == "biological and biomedical sciences" ~ "biology",
#         det_fld == "mathematics and statistics" ~ "mathematics",
#         det_fld == "physics and astronomy" ~ "physics",
#         det_fld == "business management and administration" ~ "business",
#         det_fld == "materials science engineering" ~ "materials science",
#         det_fld == "health sciences" ~ "medicine", # unsure!
#         broad_fld == "engineering" & det_fld != "materials science engineering" ~ "engineering",
#         broad_fld == "humanities and arts" & det_fld != "history" ~ "art",
#         # assign geology, geography, environmental science to geosciences in mag
#         det_fld == "geosciences, atmospheric sciences, and ocean sciences" ~ "geosciences",
#         det_fld == "agricultural sciences and natural resources" ~ "geosciences",
#         TRUE ~ "unknown"
#     ))

# # ## adjust the one geosciences 
# geo_fields <- c("geology", "geography", "environmental science")
# fields_mag <- fields_mag %>%
#     mutate(fieldname_mag = ifelse(fieldname_mag %in% geo_fields, "geosciences", fieldname_mag))

```

```{r}
d_main <- authors %>%
    mutate(has_advisor = ifelse(goid %in% advisors$goid, 1, 0),
           has_field = ifelse(goid %in% fields$goid, 1, 0),
           has_field_mag = ifelse(goid %in% fields_mag$goid, 1, 0))

fields_perauthor <- fields_mag %>%
    group_by(goid) %>%
    summarise(n_field0 = n(), .groups = "drop")

d_main <- d_main %>%
    left_join(fields_perauthor, by = "goid")

```


### Number of authors 

```{r,echo=FALSE}
author_year <- d_main %>%
    group_by(degree_year) %>%
    summarise(nb = n(), .groups = "drop")

author_year %>%
    ggplot(aes(x = degree_year, y = nb)) +
    geom_line() +
    labs(y = "Number of authors") +
    geom_vline(xintercept = 2018, linetype = "dashed")
```


### Number of fields per author 

```{r, echo=FALSE}
d_main %>%
    group_by(degree_year, has_field) %>%
    summarise(nb = n(), .groups = "drop") %>%
    group_by(degree_year) %>%
    mutate(s = nb / sum(nb)) %>%
    ungroup() %>%
    filter(has_field == 1) %>%
    ggplot(aes(x = degree_year, y = s)) +
    geom_line() +
    labs(y = "Fraction of authors with at least 1 field") +
    geom_vline(xintercept = 2018, linetype = "dashed")

```



### Advisor present per author 

- advisors are only present since 1986


```{r, echo=FALSE}
d_main %>%
    group_by(degree_year, has_advisor) %>%
    summarise(nb = n(), .groups = "drop") %>%
    group_by(degree_year) %>%
    mutate(s = nb / sum(nb)) %>%
    ungroup() %>%
    filter(has_advisor == 1) %>%
    ggplot(aes(x = degree_year, y = s)) +
    geom_line() +
    labs(y = "Fraction of authors with at least 1 advisor") +
    geom_vline(xintercept = 2018, linetype = "dashed") +
    geom_vline(xintercept = 1986, linetype = "dashed")

```


### Fraction of authors with field in MAG (overall)

```{r, echo=FALSE}
d_main %>%
    group_by(degree_year, has_field_mag) %>%
    summarise(nb = n(), .groups = "drop") %>%
    group_by(degree_year) %>%
    mutate(s = nb / sum(nb)) %>%
    ungroup() %>%
    filter(has_field_mag == 1) %>%
    ggplot(aes(x = degree_year, y = s)) +
    geom_line() +
    labs(y = "Fraction of all authors with link to MAG FoS") +
    geom_vline(xintercept = 2018, linetype = "dashed")
```

### Fraction of authors with field in MAG (as of those with at least one field)

```{r}
d_main %>%
    filter(has_field == 1) %>%
    group_by(degree_year, has_field_mag) %>%
    summarise(nb = n(), .groups = "drop") %>%
    group_by(degree_year) %>%
    mutate(s = nb / sum(nb)) %>%
    ungroup() %>%
    filter(has_field_mag == 1) %>%
    ggplot(aes(x = degree_year, y = s)) +
    geom_line() +
    labs(y = "Fraction of authors with field having link to MAG FoS") +
    geom_vline(xintercept = 2018, linetype = "dashed")
```


### Number of fields MAG level 0 per author

```{r}
d_main %>%
    filter(has_field_mag == 1 & degree_year > 1950) %>%
    mutate(n_field0 = case_when(
        n_field0 <= 3 ~ as.character(n_field0),
        n_field0 > 3 ~ "more than 3"
    )) %>%
    mutate(n_field0 = factor(n_field0)) %>%
    group_by(degree_year, n_field0) %>%
    summarise(nb = n(), .groups = "drop") %>%
    group_by(degree_year) %>%
    mutate(s = nb / sum(nb)) %>%
    ungroup() %>%
    ggplot(aes(x = degree_year, y = s, color = n_field0)) +
    geom_line() +
    labs(y = "Share") +
    theme(legend.position = "bottom")
```


### Graduates in MAG and in SED 

```{r}
# Prepare the proquest data 
nb_pq <- authors %>%
    filter(grepl("United States", location)) %>%
    inner_join(fields_mag %>% # NOTE: drops graduates w/o field in mag, most of them "education"
                filter(position == 0) %>% # there can be multiple fields per goid, need to take one
                select(goid, fieldname_mag),
            by = "goid") %>%
    rename(year = degree_year) %>%
    group_by(year, fieldname_mag) %>%
    summarise(nb = n(),
             .groups = "drop")

# assign to broad field
humanities <- c("art", "history", "philosophy")
math_compsc <- c("mathematics", "computer science")
socsci <- c("economics", "political science", "psychology", 
            "sociology")
phys_earth <- c("chemistry",
                "geography", "geology", "physics")
life_sc <- c("biology", "medicine")
engn <- c("engineering", "materials science")
unclassified <- c("environmental science", "business") # business is classified as "other social science" (see nces_graduates_field_gender.csv)
nb_pq <- nb_pq %>%
    mutate(fld = case_when(
        fieldname_mag %in% humanities ~ "humanities",
        fieldname_mag %in% math_compsc ~ "math_compsc",
        fieldname_mag %in% socsci ~ "socsci",
        fieldname_mag %in% life_sc ~ "life_sc",
        fieldname_mag %in% phys_earth ~ "phys_earth",
        fieldname_mag %in% engn ~ "engn",
        fieldname_mag %in% unclassified ~ "unclassified"
    ))

nb_pq <- nb_pq %>%
    group_by(year, fld) %>%
    summarise(nb = sum(nb),
              .groups = "drop")

```

```{r}
nb_sed <- data_graduates %>%
    group_by(year, fld) %>%
    summarise(nb = sum(nb),
              .groups = "drop")

s_theses <- nb_pq %>%
    rename(nb_pq = nb) %>%
    left_join(nb_sed %>%
                rename(nb_sed = nb),
              by = c("fld", "year")) %>%
    filter(year >= 1980) %>%
    mutate(s_pq = nb_pq / nb_sed)

n_theses <- s_theses %>%
    select(-s_pq) %>%
    gather(key = src, value = nb, nb_sed:nb_pq) %>%
    mutate(src = ifelse(src == "nb_pq", "ProQuest", "NSF")) %>%
    group_by(year, src) %>%
    summarise(nb = sum(nb, na.rm = TRUE), # nb_sed missing for unclassified
              .groups = "drop")

```

#### Totals over time

- Note: proquest has some theses classified as post-2020, they are missing from here
- In both counts, the field "education" is missing
```{r}
n_theses %>%
    filter(year <= 2019) %>%
    ggplot(aes(x = year, y = nb, linetype = src)) +
    geom_line() +
    theme(legend.position = "bottom") +
    labs(y = "Number of graduates")
```

#### Fraction ProQuest of NSF by field and year 

```{r}
s_theses %>%
    filter(!(fld %in% c("unclassified")) & year < 2020) %>%
    ggplot(aes(x = year, y = s_pq, color = fld)) +
    geom_line() +
    theme(legend.position = "bottom") +
    labs(y = "Share ProQuest of NSF") +
    geom_hline(yintercept = 1)

```


Notes:

- if an author has multiple fields, the first one is used.
- life_sc includes medicine.
- missing is the interdisciplinary field "environmental science" from ProQuest.
- Todo: repeat this with more aggregate fields: e.g. GEEMP and LPS (to the extent possible). does this get rid of the over-representation of the math/computer science fields? (for instance because of engineering?). 


### Distribution of how the links to the field were made

- Note: this is the fraction of total links per degree_year; if an author has two links, she enters twice here
```{r}
fields_mag %>%
    left_join(d_main %>%
        select(goid, degree_year),
        by = "goid") %>%
    group_by(degree_year, flag) %>%
    summarise(nb = n(), .groups = "drop") %>%
    group_by(degree_year) %>%
    mutate(s = nb / sum(nb)) %>%
    ungroup() %>%
    ggplot(aes(x = degree_year, y = s, color = flag)) +
    geom_line() +
    labs(y = "Share") +
    theme(legend.position = "bottom")
```    
    


### Questions and thoughts

- What happens around 2018? Why more information on advisor but not on field?
    - seems to be driven by non-US institutions; see comparison ProQuest-NSF
- Things of interest
    - Counts country of the institution
    - Compare counts of graduates in the US with the official US statistics
- Comparing NSF and ProQuest 
    - presumably, ProQuest covers top institutions more while NSF covers all? can we get some stats by institution as well?
    - why does the coverage seem to drop after 2010?
        - lag in data collection 
        - change in publication process?
        - change in data quality in proquest/do we drop more theses in the data cleaning process after 2010? 
        - change in reporting of university names ("United States" not in name anymore?)

```{r, include = FALSE}
DBI::dbDisconnect(con)
```
