---
title: "Duration analysis"
author: "Flavio Hafner, Christoph Hedtrich"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true 
    toc_depth: 4
---

Notes

- the output measures are re-scaled within experience cell. When using for duration analysis with time-varying covariates, check if that makes sense



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (interactive()) {
  # https://stackoverflow.com/questions/48684595/source-nested-r-files-within-rmarkdown-document. For calling files from subdir. This needs to be done only when called from within RStudio!!
  knitr_rootdir <- rprojroot::find_rstudio_root_file()
  knitr_rootdir <- paste0(knitr_rootdir, "analysis/")
  knitr::opts_knit$set(root.dir = knitr_rootdir)
}

```



## Preliminaries 

```{r, include = FALSE}
source("setup.R")
source("helpers/functions.R")
source("helpers/db_queries.R")

# Inputs for regressions
fe_controls <- c("field_f", "gradyear_f", "uni_f")
fe_controls <- paste0(fe_controls, collapse = " + ")
x_female <- "female_geemp + female_lps"
clustervar <- "field_f"

```


## Prepare data
```{r}
source("helpers/data_duration.R")
```


## Some summary stats from the data 

Rescaling the output measures 


```{r, echo = FALSE}

# is this good enough? how would we know? 
d_var <- bind_rows(
  papers_career %>% 
    mutate(y = cit_measure / ci_year_field) %>%
    group_by(Field0, Year) %>%
    summarise(v = var(y), .groups = "drop") %>%
    mutate(type = "rescale"),
  papers_career %>%
    mutate(y = cit_measure / ci_year) %>%
    group_by(Field0, Year) %>%
    summarise(v = var(y), .groups = "drop") %>%
    mutate(type = "noscale")
) 


d_var %>%
      ggplot(aes(x = log(v), group = type, color = type, fill = type)) + 
      geom_histogram(aes(y=..density..),
                     position = "identity", alpha = 0.2, bins = 60) +
      geom_density(alpha = 0.01) +
      labs(x = "Log variance within Field0-year") 


# the figure shows the plot of the field-year specific variance of citations
  # the variance slightly increases on average, but the dispersion gets much lower
  # this means that citation counts within field1 become more homogenous within field0 across years and across field1

# get similar figure for IQR 
  
```


\newpage
Career length when using Huang and exact year of graduation (for censored = 0)


```{r, echo=FALSE}
d_career %>%
  filter(censored == 0) %>%
  select(AuthorId, starts_with("career_length"), gender) %>%
  gather(key = var, value = val, career_length:career_length_huang) %>%
  ggplot(aes(val, color = gender, group = gender)) +
  geom_density() +
  facet_wrap(~var)


```



\newpage
Compare output of graduates with >= 5 and <5 years of career length


```{r, echo=FALSE}
df_kaplanmeier %>%
  select(AuthorId, eucl_idx, n_papers, gender, career_length) %>%
  mutate(sampled = ifelse(career_length >= min_careerlength, "in sample", "not in sample"),
         v_norm = pmin(career_length, min_careerlength)) %>% 
  ggplot(aes(eucl_idx, color = gender, group = gender)) +
  # geom_density() +
  stat_ecdf(geom = "step") +
  facet_wrap(~sampled)


```



## Duration analysis: Kaplan-Meier


```{r, echo=FALSE}

m_all <- survfit(Surv(career_length, status, type = "right") ~ gender, 
             ctype = 2,
             cluster = df_kaplanmeier$field_f, 
             robust = T,
             data = df_kaplanmeier)
m_lps <- survfit(Surv(career_length, status, type = "right") ~ gender, 
                 subset = df_kaplanmeier$fld == "LPS" ,
                 ctype = 2,
                 cluster = df_kaplanmeier$field_f,
                 robust = T,
                 data = df_kaplanmeier)
m_geemp <- survfit(Surv(career_length, status, type = "right") ~ gender,
                   subset = df_kaplanmeier$fld == "GEEMP",
                   ctype = 2,
                   cluster = df_kaplanmeier$field_f, 
                   robust = T,
                   data = df_kaplanmeier)


p <- autoplot(m_all) + 
  theme(legend.position = "bottom") + 
  labs(title = "All fields")

p 

ggsave(plot = p, 
       filename = paste0(figdir, "career_kaplanmeier.pdf"),
       device = "pdf")


autoplot(m_lps) + 
  theme(legend.position = "bottom") + 
  labs(title = "Field: LPS")

autoplot(m_geemp) + 
  theme(legend.position = "bottom") + 
  labs(title = "Field: GEEMP") 


```




## Gender differences in output and citations

### Figures: euclidean index of citations


CDF

```{r}
p <- df_returns %>%
  ggplot(aes(x = eucl_idx, group = gender, color = gender)) +
  stat_ecdf() +
  theme(legend.position = "bottom") + 
  labs(y = "Percentile", x = "Euclidean citation index")

ggsave(plot = p, 
       filename = paste0(figdir, "eucl_idx_ecdf.pdf"),
       device = "pdf")

p

```

\newpage
Average index along percentiles of paper count 

```{r}
probs_q <- c(0, 0.25, 0.5, 0.6, 0.7, 0.8, 0.9, 1)
# probs_q <- seq(0, 1, 0.25)
df_returns %>% 
  mutate(grp = cut(n_papers, 
                   breaks = quantile(n_papers, probs = probs_q), 
                   labels = probs_q[2:length(probs_q)],
                   include.lowest = T)
         ) %>%
  mutate(grp = as.numeric(as.character(grp))) %>%
  group_by(grp, gender) %>%
  summarise(y_qual = mean(eucl_idx),
            .groups = "drop") %>%
  ggplot(aes(x = grp, y = y_qual, color = gender, group = gender)) +
  geom_line() +
  geom_point() +
  labs(x = "Percentile(#Papers)", y = "Average Euclidean index") 

# no big differences between fields. 
```


Paper count by field 

```{r}

df_returns %>%
  group_by(gender, main_field_f) %>%
  summarise(n_papers = mean(n_papers_lvl),
            .groups = "drop") %>%
  spread(key = gender, value = n_papers) %>%
  mutate(y = (Female - Male) / Male) %>%
  ggplot(aes(x = main_field_f, y = y)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  coord_flip()

```



### Regressions: average 

- number of obs may change b/c all outcomes are 0 within a group.

```{r}

df_returns <- df_returns %>%
  mutate(post95 = ifelse(degree_year >= 1995, 1, 0),
         time_trend = degree_year - 1995,
         n_papers2 = n_papers * n_papers,
         eucl_idx2 = eucl_idx * eucl_idx,
         eucl_idx_papers = eucl_idx * n_papers,
         avg_cites_lvl = n_cites_lvl / n_papers_lvl
         ) %>%
  mutate(female_geemp = ifelse(fld == "GEEMP" & female == 1, 1, 0),
         female_lps = ifelse(fld == "LPS" & female == 1, 1, 0))

ls_specs <- list(
  papers = paste0("n_papers_lvl ~ ", x_female),
  cites = paste0("n_cites_lvl ~ ", x_female),
  cites_p = paste0("n_cites_lvl ~ ", x_female, "+ n_papers_lvl"),
  avg_cites = paste0("avg_cites_lvl ~ ", x_female)
)



ls_models <- map(
  .x = ls_specs,
  .f = ~fepois(
    fml = as.formula(paste0(.x,
                            "| ",
                            fe_controls
                            )
                     ),
    nthreads = 1,
    cluster = clustervar,
    data = df_returns
  )
)

ls_specs_idx <- list(
  idx = paste0("eucl_idx ~ ", x_female),
  idx_p = paste0("eucl_idx ~ ", x_female, "+ n_papers")
)

ls_models_idx <- map(
  .x = ls_specs_idx,
  .f = ~feols(
    fml = as.formula(paste0(.x,
                            "| ",
                            fe_controls
                            )
                     ),
    nthreads = 1,
    cluster = clustervar,
    data = df_returns
  )
)


ls_out<- c(ls_models, ls_models_idx)


etable(ls_out,
     drop = c("BIC"),
     sdBelow = T,
     convergence = TRUE,
     fitstat = c("n", "pr2")
)

etable(ls_out,
     drop = c("BIC"),
     sdBelow = T,
     convergence = FALSE,
     file = paste0(tabledir, "gap_pub_avg.tex"),
     digits = 2,
     replace = TRUE,
     signifCode=NULL,
     fitstat = c("n", "pr2")
)



```


### Regressions: quartiles. use multinomial model instead? 

```{r}

lapply(c("qnt_papers", "qnt_idx"), function(yvar) {
  
  d_temp <- df_returns %>%
    mutate(y_q4 = ifelse(.data[[yvar]] == 4, 1, 0),
           y_q3 = ifelse(.data[[yvar]] == 3, 1, 0),
           y_q2 = ifelse(.data[[yvar]] == 2, 1, 0),
           y_q1 = ifelse(.data[[yvar]] == 1, 1, 0))

  ls_specs <- paste0("y_q", 1:4)
  ls_specs <- map(.x = ls_specs,
                  .f = ~paste0(.x, "~ ", x_female))
  
  ls_models <- map(
    .x = ls_specs,
    .f = ~feglm(
      fml = as.formula(paste0(.x,
                              "| ", 
                              fe_controls
                              )
                       ),
      nthreads = 1,
      family = "binomial",
      cluster = clustervar,
      data = d_temp
    )
  )
  
  
  etable(ls_models,
     drop = c("BIC"),
     sdBelow = T,
     convergence = FALSE,
     file = paste0(tabledir, "gap_", yvar, ".tex"),
     replace = TRUE,
     digits = 2,
     signifCode=NULL,
     fitstat = c("n", "pr2")
  )
  
  etable(ls_models,
         title = yvar,
         drop = c("BIC"),
         sdBelow = T,
         convergence = TRUE,
         fitstat = c("n", "pr2")
  )


})

```



## Covariance quality-quantity?


### Decomposition

- the size of the diamonds is the number of graduates in the field in our sample 


```{r}
sel_measures <- c("CitationPerPaper", "CovQualQuant", "PaperCount")

plot_decomp <- df_decomp %>% 
  filter(measure %in% sel_measures) %>%
  mutate(main_field_f = fct_reorder(main_field_f, gap_total)) %>%
  filter(exper <= 14)


ggplot() +
  geom_bar(data = plot_decomp,
           aes(x = main_field_f, 
               y = gap, 
               fill = measure),
           stat = "identity") +
  geom_line(data = plot_decomp,
            aes(x = main_field_f,
                y = gap_total)) +
  geom_point(data = plot_decomp,
             aes(x = main_field_f,
                 y = gap_total, 
                 size = AuthorCount_Total),
             shape = 18) +
  theme(legend.position = "bottom") +
  coord_flip() +
  facet_wrap(~exper) +
  labs(y = "Gap", x = "Field") +
  guides(size = "none", color = "none")


```


## Link to PhD supervision data 


```{r}


d_advisor <- advisor_links %>%
  group_by(AuthorId) %>%
  filter(advisee_degree_year == min(advisee_degree_year)) %>%
  ungroup()

n_students <- advisor_links %>%
  group_by(AuthorId) %>%
  summarise(nb = n(),
            .groups = "drop")


d_advisor_all <- df_kaplanmeier %>%
  left_join(d_advisor %>%
              select(AuthorId, first_advisor_year = advisee_degree_year),
            by = "AuthorId") %>%
  mutate(advisor = ifelse(is.na(first_advisor_year), 0, 1),
         dur_advisor = ifelse(
           advisor == 1, first_advisor_year - degree_year,
           2015 - degree_year
         )) %>%
  filter(dur_advisor > 0)
  
d_advisor <- df_returns %>%
  left_join(d_advisor %>%
              select(AuthorId, first_advisor_year = advisee_degree_year),
            by = "AuthorId")  %>%
  mutate(advisor = ifelse(is.na(first_advisor_year), 0, 1),
         dur_advisor = ifelse(
           advisor == 1, first_advisor_year - degree_year,
           2015 - degree_year
         )) %>%
  filter(dur_advisor > 0) %>%
  left_join(n_students %>%
              rename(n_students = nb),
            by = "AuthorId") %>%
  mutate(n_students = ifelse(is.na(n_students), 0, n_students))
  
d_advisor_all <- d_advisor_all %>%
  filter(dur_advisor < (2015-1985))
d_advisor <- d_advisor %>%
  filter(dur_advisor < (2015-1985))

```


### Descriptives: what does advisor measure? Add: average publication tenure at the (first?) advising institution! for censored and not censored?


```{r}
m0 <- feols(career_length ~ advisor,
           data = dk,
           cluster = clustervar)

m1 <- feols(career_length ~ advisor | gradyear_f + uni_f + field_f,
           data = dk,
           cluster = clustervar)

etable(list(m0, m1),
     drop = c("BIC"),
     sdBelow = T,
     convergence = TRUE,
     fitstat = c("n", "pr2")
)

```



### Outcome: becoming an advisor


Figure 

```{r}

d_advisor_all %>% 
  mutate(degree_year = 5 * floor(degree_year / 5)) %>%
  group_by(fld, gender, degree_year) %>%
  summarise(advisor = mean(advisor), .groups = "drop") %>%
  ggplot(aes(x = degree_year, y = advisor, group = gender, color = gender)) +
  geom_line() + 
  facet_wrap(~fld)

p <- d_advisor_all %>% 
  mutate(year = ifelse(degree_year < 1995, "1985-1994", "1995-2005")) %>%
  group_by(year, fld, gender) %>%
  summarise(advisor_mn = mean(advisor), 
            advisor_var = var(advisor),
            N = n(),
            .groups = "drop") %>%
  mutate(se = sqrt(advisor_var / N),
         ci_ub = advisor_mn + 1.96 * se,
         ci_lb = advisor_mn - 1.96 * se) %>%
  ggplot(aes(x = year, y = advisor_mn, group = gender, fill = gender)) +
  geom_bar(stat = "identity", position = position_dodge(0.9)) +
  geom_errorbar(position = position_dodge(0.9),
                aes(ymin = ci_lb, ymax = ci_ub),
                width = 0.2) +
  facet_wrap(~fld) +
  theme(legend.position = "bottom") + 
  labs(y = "P(becomes advisor)")

p 

ggsave(plot = p, filename = paste0(figdir, "prob_advisor.pdf"),
       device = "pdf")


```

Summary statistics

```{r}

vars_sumstat <- c("degree_year", "link_score", "female", "career_length", 
                  "n_cites_lvl", "n_papers_lvl", "eucl_idx_lvl", "advisor")
sum_stats <- c("mean", "median", "sd", "min", "max")

tb <- capture.output(
  stargazer(as.data.frame(d_advisor[ , vars_sumstat]), 
          summary = T, digits = 2, digit.separator = "",
          summary.stat = sum_stats)
)


tb <- tb[!grepl("table", tb)]
writeLines(tb, paste0(tabledir, "sumstat_returns.tex"))


```


```{r}

p <- d_advisor %>%
  group_by(main_field_f, gender) %>%
  summarise(nb = n(), .groups = "drop") %>%
  group_by(main_field_f) %>%
  mutate(tot = sum(nb)) %>%
  ungroup() %>%
  mutate(main_field_f = fct_reorder(main_field_f, tot)) %>%
  ggplot(aes(x = main_field_f, y = nb, fill = gender, group = gender)) +
  geom_bar(stat = "identity") +
  theme(legend.position = "bottom") + 
  labs(y = "Count", x = "Field") +
  coord_flip()

ggsave(plot = p,
       filename = paste0(figdir, "sample_count_fields.pdf"),
       device = "pdf")

p

```




Regressions: Gap in prob(advisor)

```{r}


ls_specs <- list(
  base =  paste0(x_female, "| gradyear_f"),
  fld =  paste0(x_female, "| gradyear_f + field_f"),
  ctrl = paste0(x_female, "|", fe_controls),
  papers = paste0(x_female, " + n_papers", "|", fe_controls),
  cites = paste0(x_female, " + n_cites", "|", fe_controls),
  idx = paste0(x_female, " + eucl_idx", "|", fe_controls),
  idx_p = paste0(x_female, " + eucl_idx + n_papers", "|", fe_controls)
)


ls_models <- map(
  .x = ls_specs,
  .f = ~feglm(
    fml = as.formula(paste0("advisor ~ ", .x
                            )
                     ),
    nthreads = 1,
    cluster = clustervar,
    family = "binomial",
    data = d_advisor
  )
)


etable(ls_models,
     drop = c("BIC", "Intercept"),
     sdBelow = T,
     convergence = TRUE,
     fitstat = c("n", "pr2")
)

etable(ls_models,
     drop = c("BIC", "Intercept"),
     sdBelow = T,
     convergence = FALSE,
     digits = 2,
     signifCode=NULL,
     file = paste0(tabledir, "gap_advisor.tex"),
     replace = TRUE,
     fitstat = c("n", "pr2")
)


```


```{r}

x_female_time <- "female_geemp*post95 + female_lps*post95"

ls_specs <- list(
  base = x_female_time,
  ctrl = paste0(x_female_time, "|", fe_controls),
  papers = paste0(x_female_time, " + n_papers*post95", "|", fe_controls),
  papers_qrt = paste0(x_female_time, " + factor(qnt_papers)*post95", "|", fe_controls)
)


ls_models <- map(
  .x = ls_specs,
  .f = ~feglm(
    fml = as.formula(paste0("advisor ~ ", .x
                            )
                     ),
    nthreads = 1,
    cluster = clustervar,
    family = "binomial",
    data = d_advisor
  )
)


etable(ls_models,
     drop = c("BIC", "Intercept", "n_papers", "qnt_papers"),
     sdBelow = T,
     convergence = TRUE,
     fitstat = c("n", "pr2")
)

etable(ls_models,
     drop = c("BIC", "Intercept", "n_papers", "qnt_papers"),
     sdBelow = T,
     convergence = FALSE,
     digits = 2,
     signifCode=NULL,
     file = paste0(tabledir, "gap_advisor_time.tex"),
     replace = TRUE,
     fitstat = c("n", "pr2")
)


```



### Returns to producing more output for becoming an advisor 

```{r}
x_female_all <- "female"


ls_specs <- list(
  papers = paste0(x_female_all, " + n_papers * female ", "|", fe_controls),
  idx = paste0(x_female_all, " + eucl_idx * female ", "|", fe_controls),
  avg_cites = paste0(x_female_all, " + avg_cites * female ", "|", fe_controls),
  idx_p = paste0(x_female_all, " + eucl_idx * female  + n_papers * female ", "|", fe_controls),
  avg_cites_p = paste0(x_female_all, " + avg_cites * female  + n_papers * female ", "|", fe_controls)
)


lapply(c("LPS", "GEEMP"), function(fld) {
  ls_models <- map(
  .x = ls_specs,
  .f = ~feglm(
    fml = as.formula(paste0("advisor ~ ", .x
                            )
                     ),
    nthreads = 1,
    cluster = clustervar,
    family = "binomial",
    data = d_advisor[d_advisor$fld == fld, ]
  )
  )
  
  etable(ls_models,
       drop = c("BIC", "Intercept"),
       sdBelow = T,
       convergence = FALSE,
       digits = 2,
       signifCode=NULL,
       file = paste0(tabledir, "gap_returns_", tolower(fld), ".tex"),
       replace = TRUE,
       fitstat = c("n", "pr2")
  )
  
  etable(ls_models,
       drop = c("BIC", "Intercept"),
       sdBelow = T,
       convergence = TRUE,
       fitstat = c("n", "pr2")
  )
    
  
})

```



```{r}
DBI::dbDisconnect(con)
```
