---
title: "Graduation trends in proquest and official statistics"
author: "Flavio Hafner, Christoph Hedtrich"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true 
    toc_depth: 4
---


Questions:

1. Is gender assignment of graduates stable? 
2. Does our transition rate from PhD to publishing correlate with the official transition rate?



To do 

- tidy script below 
- integrate with `quality_current_links.Rmd`?
- merge this with quality_current_links? it seems to have mostly the same purpose. not sure what is a good name for it 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (interactive()) {
  # https://stackoverflow.com/questions/48684595/source-nested-r-files-within-rmarkdown-document. For calling files from subdir. This needs to be done only when called from within RStudio!!
  knitr_rootdir <- rprojroot::find_rstudio_root_file()
  knitr_rootdir <- paste0(knitr_rootdir, "analysis/")
  knitr::opts_knit$set(root.dir = knitr_rootdir)
}

```


## Preliminaries 

```{r, include = FALSE}
source("setup.R")
source("helpers/functions.R")

path_nces <- "/mnt/ssd/NCES_NSF/processed/"
figdir <- "../../img/duration_advisor/"

```


Notes 

- the identification of links will have some measurement error because some authors are identified in different fields than field0


```{r}

# what do I need?
q_authors <- "
SELECT * from pq_authors
INNER JOIN (
  SELECT university_id, normalizedname AS university_name 
  FROM pq_unis
  WHERE location like '%United States%'
) USING(university_id)
"

pq_authors <- tbl(con, sql(q_authors))


pq_fields_mag <- tbl(con, "pq_fields_mag") %>% # see the quality_proquest file for a correspondence? 
  filter(position == 0) %>%
  select(goid, mag_field0) %>%
  left_join(tbl(con, "FieldsOfStudy") %>%
              select(FieldOfStudyId, fieldname = NormalizedName),
            by = c("mag_field0" = "FieldOfStudyId"))


firstnames_gender <- tbl(con, "FirstNamesGender") %>%
  mutate(gender = case_when(
    ProbabilityFemale <= 0.2 ~ "male",
    ProbabilityFemale >= 0.8 ~ "female"
  ))
current_links <- tbl(con, "current_links")

affiliations <- tbl(con, "Affiliations") %>%
  select(AffiliationId, NormalizedName, Iso3166Code)

author_affiliation <- tbl(con, "AuthorAffiliation") %>%
  inner_join(current_links %>%
               select(AuthorId),
             by = "AuthorId") %>%
  left_join(affiliations,
            by = "AffiliationId") %>%
  select(AuthorId, Year, AffiliationName = NormalizedName, Iso3166Code)

```



```{r}
counts_nces <- read.csv(paste0(path_nces, "graduate_counts.csv"))
fields_nces <- read.csv(paste0(path_nces, "fields.csv"))

parent_fields_nces <- read.csv(paste0(path_nces, "parent_fields.csv"))

counts_us_nces <- read.csv(paste0(path_nces, "postgrad_us_counts.csv"))
sectors_us_nces <- read.csv(paste0(path_nces, "postgrad_us_sectors.csv"))

counts_nces <- counts_nces %>%
  left_join(fields_nces %>%
              select(field_id, fld = shortname),
            by = "field_id")

acad_us_nces <- sectors_us_nces %>%
  filter(sector == "academia") %>%
  left_join(counts_us_nces, 
            by = c("gender", "field_id", "year")) %>%
  mutate(n_acad = share * nb) %>%
  select(field_id, year, gender, n_acad)
# aggregate by major field 
acad_us_nces <- acad_us_nces %>%
  left_join(parent_fields_nces, by = "field_id") %>%
  group_by(year, parent_field_id, gender) %>%
  summarise(n_acad = sum(n_acad), .groups = "drop")

acad_us_nces <- acad_us_nces %>%
  left_join(counts_nces %>%
              select(field_id, year, gender, nb),
            by = c("parent_field_id" = "field_id", "year", "gender")) %>%
  rename(field_id = parent_field_id, 
         n_graduates = nb) %>%
  left_join(fields_nces %>%
              select(field_id, fld = shortname),
            by = "field_id")


```


```{r}

pq_authors <- pq_authors %>%
  filter(degree_year >= 1980 & degree_year <= 2021) %>%
  select(goid, degree_year, firstname, degree_year, university_name) %>%
  left_join(firstnames_gender %>%
              select(FirstName_g = FirstName, gender),
            by = c("firstname" = "FirstName_g")) %>%
  left_join(pq_fields_mag, by = "goid") %>%
  collect()

```

```{r}
humanities <- c("art", "history", "philosophy")
math_compsc <- c("mathematics", "computer science")
socsci <- c("economics", "political science", "psychology", 
            "sociology")
phys_earth <- c("chemistry", "environmental science", # env sc seems to make sense to be put here 
                "geography", "geology", "physics")
life_sc <- c("biology", "medicine")
engn <- c("engineering", "materials science")
unclassified <- c("environmental science", "business") # business is classified as "other social science" (see nces_graduates_field_gender.csv)
unclassified <- c("business")
pq_authors <- pq_authors %>%
    mutate(fld = case_when(
        fieldname %in% humanities ~ "humanities",
        fieldname %in% math_compsc ~ "math_compsc",
        fieldname %in% socsci ~ "socsci",
        fieldname %in% life_sc ~ "life_sc",
        fieldname %in% phys_earth ~ "phys_earth",
        fieldname %in% engn ~ "engn",
        fieldname %in% unclassified ~ "unclassified",
        TRUE ~ "missing"
    ))


```


## 1. Is gender assignment stable?

Compare gender trends by field in proquest and in official data

```{r}
select_fields <- c("engn", "humanities", "life_sc", "math_compsc", "phys_earth",
                   "socsci")

counts_pq <- pq_authors %>% 
  group_by(degree_year, gender, fld) %>%
  summarise(nb = n(), .groups = "drop") %>%
  group_by(degree_year, fld) %>%
  mutate(s = nb / sum(nb)) %>%
  ungroup()

counts_nces <- counts_nces %>%
  group_by(year, fld) %>%
  mutate(s = nb / sum(nb)) %>%
  ungroup()

d_counts <- bind_rows(
  counts_pq %>% 
    select(gender, fld, year = degree_year, share = s, nb) %>%
    mutate(source = "ProQuest"),
  counts_nces %>%
    select(gender, fld, year, share = s, nb) %>%
    mutate(source = "NCES")
) %>%
  filter(year <= 2019) %>%
  filter(fld %in% select_fields) %>%
  mutate(gender = ifelse(is.na(gender), "missing", gender)) 


```



```{r}

p <- d_counts %>% 
  filter(fld != "humanities") %>%
  group_by(fld, year, source) %>%
  summarise(nb = sum(nb),
            .groups = "drop") %>%
  ggplot(aes(x = year, y = nb, linetype = source)) +
  geom_line() +
  facet_wrap(~fld) + 
  theme(legend.position = "bottom") +
  annotate("rect", xmin = 1985, xmax = 2005, ymin = 0, ymax = 16000, 
           alpha = .2) +
  scale_x_continuous(breaks = c(1980, 1990, 2000, 2010))

p 

ggsave(plot = p,
       filename = paste0(figdir, "trends_pq_nces.pdf"),
       device = "pdf")

p <- d_counts %>% 
  filter(fld != "humanities") %>%
  filter(gender %in% c("male", "female")) %>%
  ggplot(aes(x = year, y = nb, color = gender, linetype = source)) +
  geom_line() +
  facet_wrap(~fld) + 
  theme(legend.position = "bottom") +
  annotate("rect", xmin = 1985, xmax = 2005, ymin = 0, ymax = 8000, 
           alpha = .2) +
  scale_x_continuous(breaks = c(1980, 1990, 2000, 2010))

p 

ggsave(plot = p,
       filename = paste0(figdir, "trends_pq_nces_gender.pdf"),
       device = "pdf")

```

Observations

- Time trends are roughly similar. But in some fields the opening difference for women is large (double in maths/CS for instance)

- Constant difference in coverage for men; opening coverage for women over time 

- if it was driven by asians where gender are less well assigned, there should be an opening difference for men and women 

- Instead, it must be female-specific. Either more Asian women, or the American women have more gender-"neutral" names since 1990/2000

- why is engineering undercovered? Does not seem to be from compsc since the coverage is almost 100 percent there 

- undercoverage in life sciences: why? 


## 2. Does our transition rate from PhD to publishing correlate with the official transition rate?

Transitions to US academia in NCES

```{r}
author_affiliation <- collect(author_affiliation)
current_links <- collect(current_links)

```


```{r}
first_affil_us <- author_affiliation %>%
  # drop affiliations pre-Phd
  left_join(current_links %>%
              select(goid, AuthorId),
            by = "AuthorId") %>%
  left_join(pq_authors %>%
              select(goid, degree_year, university_name ),
            by = "goid") %>%
  filter(Year > degree_year) %>%
  group_by(AuthorId) %>%
  filter(Year == min(Year)) %>%
  ungroup() %>%
  filter(Iso3166Code == "US") %>%
  filter(!duplicated(AuthorId)) %>%
  mutate(dist_uni_names = stringdist(
    AffiliationName, university_name, method = "jw"
  )) %>%
  mutate(same_uni = ifelse(dist_uni_names > 0, "no", "yes"))
#NOTE: unclear where to cut, even with distance of 0.05 it is still possible to be the same uni ("at")
  # but also univ california system will have such small distances
```


```{r}

pq_to_us_acad <- pq_authors %>%
  inner_join(current_links %>%
               select(AuthorId, goid),
             by = "goid") %>%
  inner_join(first_affil_us %>%
               select(AuthorId, AffiliationName, affil_year = Year, same_uni),
             by = "AuthorId") %>%
  mutate(ydiff = affil_year - degree_year) %>%
  filter(abs(ydiff) <= 5) 

uni_sector <- "university|college|institute"
affils_fld <- pq_to_us_acad %>%
  filter(degree_year %in% 1985:2005) %>%
  mutate(institution = case_when(
    same_uni == "yes" ~ "same as graduation",
    grepl(uni_sector, AffiliationName) ~ "other than graduation",
    TRUE ~ "non-academic"
  )) %>%
  group_by(fld, institution) %>%
  summarise(nb = n(), .groups = "drop")

```



Destinations of identified links in Proquest

```{r}

affils_fld %>%
  filter(!(fld %in% c("unclassified", "humanities", "missing"))) %>%
  ggplot(aes(x = fld, y = nb, fill = institution)) +
  geom_bar(stat = "identity") + 
  coord_flip() +
  theme(legend.position = "bottom")

```

Transitions to academic sector: total

```{r}

pq_to_us_acad_total <- pq_to_us_acad %>%
  filter(grepl(uni_sector, AffiliationName)) %>%
  group_by(degree_year, gender, fld) %>%
  summarise(n_acad = n(), .groups = "drop")

acad_us <- bind_rows(
  acad_us_nces %>%
    select(-field_id, -n_graduates) %>%
    mutate(source = "NCES"),
  pq_to_us_acad_total %>%
    rename(year = degree_year) %>%
    mutate(source = "ProQuest")
) %>%
  filter(year <= 2019) %>%
  filter(fld %in% select_fields) %>%
  mutate(gender = ifelse(is.na(gender), "missing", gender)) 


```


```{r}
p <- acad_us %>% 
  filter(fld != "humanities" & year %in% 1985:2005) %>%
  filter(gender %in% c("male", "female")) %>%
  ggplot(aes(x = year, y = n_acad, color = gender, linetype = source)) +
  geom_line() +
  facet_wrap(~fld) + 
  theme(legend.position = "bottom") +
  scale_x_continuous(breaks = c(1990, 2000))

ggsave(plot = p,
       filename = paste0(figdir, "transitions_pq_nces.pdf"),
       device = "pdf")
p
```

```{r}
p <- acad_us_nces %>%
  filter(fld != "humanities" & fld != "other" & year %in% 1985:2005) %>%
    ggplot(aes(x = year, y = n_acad / n_graduates, color = gender)) + 
  geom_line() + 
  facet_wrap(~fld)  +
  scale_x_continuous(breaks = c(1990, 2000)) + 
  theme(legend.position = "bottom") + 
  labs(y = "Transition rate to US academia")

ggsave(plot = p,
       filename = paste0(figdir, "transition_rates_nces.pdf"),
       device = "pdf")
p

```



Transitions to academic sector: by same/different institution than PhD institution

```{r}

pq_to_us_acad_inst <- pq_to_us_acad %>%
  filter(grepl(uni_sector, AffiliationName)) %>%
  group_by(degree_year, gender, fld, same_uni) %>%
  summarise(n_acad = n(), .groups = "drop") %>% 
  rename(year = degree_year) %>%
  filter(fld %in% select_fields & fld != "humanities") %>%
  filter(!is.na(gender)) %>%
  mutate(institution = ifelse(same_uni == "yes", "same as PhD", "other than PhD"),
         institution = factor(institution, levels = c("same as PhD", "other than PhD"))
         )

acad_us_nces <- acad_us_nces %>%
  filter(year <= 2019 & fld != "humanities" ) %>%
  filter(fld %in% select_fields) %>%
  mutate(gender = ifelse(is.na(gender), "missing", gender)) 

```



```{r}

p <- ggplot() + 
  geom_line(data = acad_us_nces %>% 
              filter(year %in% 1985:2005),
            aes(x = year, y = n_acad)) + 
  geom_bar(data = pq_to_us_acad_inst,
           aes(x = year, 
               y = n_acad, 
               fill = institution),
           alpha = 0.5,
           stat = "identity") +
  facet_grid(gender~fld) +
  theme(legend.position = "bottom") + 
  scale_x_continuous(breaks = c(1990, 2000))


ggsave(plot = p,
       filename = paste0(figdir, "transitions_pq_detailed.pdf"))

p

```



### Comments 

- black line is the NCES count (add legend?)
- the red bars are people whose first affiliation after PhD is their graduating institution. consists of
  - stayers that do postdocs at their home institution
  - stayers that publish PhD papers with a lag
  - quitters that publish PhD papers with a lag


```{r}
DBI::dbDisconnect(con)
```
