---
title: "Some stats on how well current links perform"
author: "Flavio Hafner, Christoph Hedtrich"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    toc: true 
    toc_depth: 4
---


Note: This could possibly be merged with /src/quality_linking.Rmd. But for now, I have left the two separate. Here we have some more sample restrictions and only use one sample of links. 
This is not yet recorded in MAG. 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (interactive()) {
  # https://stackoverflow.com/questions/48684595/source-nested-r-files-within-rmarkdown-document. For calling files from subdir. This needs to be done only when called from within RStudio!!
  knitr_rootdir <- rprojroot::find_rstudio_root_file()
  knitr_rootdir <- paste0(knitr_rootdir, "analysis/")
  knitr::opts_knit$set(root.dir = knitr_rootdir)
}

```



## Preliminaries 

```{r, include = FALSE}
source("setup.R")
source("helpers/functions.R")
source("helpers/db_queries.R")

figdir <- "../../img/duration_advisor/"

```


## Load relevant data 
```{r}


# this is taken from data_duration.R. Need to think how to best combine the calls without redundancies
d_career <- tbl(con, "pq_authors") %>%
  select(goid, degree_year, university_id) %>% 
  inner_join(current_links, by = "goid") %>%
  inner_join(author_sample %>%
               select(AuthorId, YearLastPub, YearFirstPub),
             by = c("AuthorId")) %>%
  inner_join(author_info %>%
               select(AuthorId, gender, censored), 
             by = "AuthorId") %>%
  filter(degree_year - YearFirstPub >= -5
         & degree_year - YearFirstPub <= 5)  %>%
  filter(degree_year >= 1985 & degree_year <= 2005) %>% 
  collect()

author_fields <- collect(author_fields)
graduates_fields <- collect(graduates_fields)
fields_names <- collect(fields_names)
current_links <- collect(current_links)


```


## Comparing the links between the graduating and the publishing field

```{r}
# first_field_id is field1 in MAG. main_field is the major field

compare_field_links <- graduates_fields %>%
  inner_join(author_fields,
             by = c("AuthorId", "field_name" = "main_field")) %>%
  select(goid, position, name_field0 = field_name,
         field1_pq = fieldname, first_field_id, field_group) %>%
  inner_join(fields_names %>%
               rename(field1_mag = NormalizedName),
             by = c("first_field_id" = "FieldOfStudyId")) %>%
  filter(goid %in% d_career$goid)
  


```


- How to read the figure: darker color = a larger fraction of graduates in graduating field (y-axis) is matched to entities in publishing field on x-axis.


```{r}


dtemp <- compare_field_links %>%
    group_by(name_field0, field1_mag, field1_pq) %>%
    summarise(nb = n(),
                .groups = "drop") %>%
    group_by(name_field0, field1_pq) %>%
    mutate(s = nb / sum(nb))

fields <- unique(dtemp$name_field0)
plots <- map(.x = fields,
    .f = ~dtemp %>% 
      filter(name_field0 == .x) %>%
      ggplot(aes(x = field1_mag, y = field1_pq)) + 
      geom_tile(aes(fill = s)) +
      scale_fill_gradient2(low = "red", mid = "yellow", high = "blue",
                           midpoint = quantile(dtemp$s, 0.5)) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
      labs(title = paste0("Major field: ", .x),
           y = "Graduating field", x = "Publishing field")
      )
names(plots) <- fields

plots

fields_to_save <- c("biology", "physics", "chemistry")

lapply(fields_to_save, function(f) {
  ggsave(plot = plots[[f]],
         filename = paste0(figdir, "fieldlink_", f, ".pdf"),
         device = "pdf")
})


```



## Fraction linked by year and field 


```{r}

pq_authors <- tbl(con, "pq_authors") %>%
  left_join(tbl(con, "pq_unis") %>%
              select(location, university_id),
            by = "university_id") %>%
  select(goid, fullname, degree_year, location) %>%
  filter(degree_year >= 1985 & degree_year <= 2005)


pq_fields = tbl(con, "pq_fields")  %>%
  inner_join(pq_authors %>%
               select(goid),
             by = "goid")
pq_fields_mag <- tbl(con, "pq_fields_mag") %>%
  inner_join(pq_authors %>%
               select(goid),
             by = "goid")


# note: for the analysis, I used the field in mag, but here this is not possible. see the linking script and/or the other quality file for code 

```



```{r}
pq_authors <- collect(pq_authors)
pq_fields <- collect(pq_fields)
pq_fields_mag <- collect(pq_fields_mag)

# keep only the first mentioned field per author that maps into MAG
pq_fields_mag <- pq_fields_mag %>%
              group_by(goid) %>%
              filter(position == min(position)) %>%
              ungroup() 

```


```{r}
d_authors <- pq_authors %>%
  filter(grepl("United States", location)) %>%
  left_join(pq_fields_mag %>%
              select(goid, mag_field0),
            by = "goid") %>%
  left_join(fields_names %>%
              rename(field = NormalizedName), 
            by = c("mag_field0" = "FieldOfStudyId")) %>%
  mutate(has_field = ifelse(goid %in% pq_fields$goid, 1, 0))

d_authors %>% 
  mutate(has_field_mag = ifelse(is.na(mag_field0), 0, 1)) %>%
  group_by(degree_year, has_field_mag) %>%
  summarise(nb = n(), .groups = "drop") %>%
  group_by(degree_year) %>%
  mutate(s = nb / sum(nb)) %>%
  ungroup() %>%
  filter(has_field_mag == 1) %>%
  ggplot(aes(x = degree_year, y = s)) +
  geom_line() + 
  labs(y = "Fraction of graduates with a link to MAG")


grad_counts <- d_authors %>%
  filter(!is.na(field)) %>%
  group_by(degree_year, field) %>%
  summarise(nb = n(), .groups = "drop") %>%
  filter(field %in% c(geemp_fields, lps_fields))

keep_links <- current_links %>%
  filter(link_score > 0.7)

link_counts <- compare_field_links %>%
  filter(goid %in% keep_links$goid) %>%
  left_join(d_authors %>%
              select(goid, degree_year),
            by = "goid") %>%
  rename(field = name_field0) %>%
  group_by(degree_year, field) %>%
  summarise(nb = n(), .groups = "drop")
  


map(.x = list(geemp_fields, lps_fields),
    .f = ~grad_counts %>%
          filter(field %in% .x) %>%
          ggplot(aes(x = degree_year, y = nb, color = field)) +
          geom_line() +
          labs(y = "Number of graduates")
  )


# these are the graduates with fields--mostly teaching degrees
# missing_mag <- dk %>% 
#   filter(is.na(mag_field0)) %>%
#   left_join(pq_fields %>%
#               filter(position == 0) %>%
#               select(goid, fieldname),
#             by = "goid")

# So: when linking field x, we use any graduates who have mag field x. 
  # this drop those without field. but it is also possible that the same person is in two fields
  # this also needs to be considered when using the links for the heatmaps above?

```



```{r}


map(.x = list(geemp_fields, lps_fields),
    .f = ~grad_counts %>%
          filter(field %in% .x) %>%
          left_join(link_counts %>% 
                      rename(n_linked = nb), 
                    by = c("degree_year", "field")) %>%
          mutate(share_linked = n_linked / nb) %>%
          ggplot(aes(x = degree_year, y = share_linked, color = field)) +
          geom_line() +
          labs(y = "Fraction linked")
  )

# Note: fraction linked 
  # the SED data are on the US, not overall. So it is difficult to compare 
  # perhaps at least compare the total numbers? but that is already in quality_proquest?

```



```{r, include = FALSE}
DBI::dbDisconnect(con)
```




